---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "tools/README-"
)
```

![stability-unstable](https://img.shields.io/badge/stability-unstable-yellow.svg)
[![Travis-CI Build Status](https://travis-ci.org/poissonconsulting/mbr.svg?branch=master)](https://travis-ci.org/poissonconsulting/mbr)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/poissonconsulting/mbr?branch=master&svg=true)](https://ci.appveyor.com/project/poissonconsulting/mbr)
[![Coverage Status](https://img.shields.io/codecov/c/github/poissonconsulting/mbr/master.svg)](https://codecov.io/github/poissonconsulting/mbr?branch=master)
[![License: MIT](https://img.shields.io/badge/License-MIT-blue.svg)](https://opensource.org/licenses/MIT)

# mbr

## Introduction

`mbr` (pronounced MBR) is an R package of utility functions and virtual classes for the model builder family of R packages.

Currently there are packages which implement the functionality for JAGS ([jmbr](https://github.com/poissonconsulting/jmbr)) and TMB ([tmbr](https://github.com/poissonconsulting/tmbr)).

`mbr` implements some basic functionality using `lm`.

## Demonstration

```{r, message = FALSE}
library(magrittr)
library(ggplot2)
library(mbr)
```

```{r}
model <- model("Pairs ~ poly(Year, 3)")

model %<>% update_model(
  select_data = list("Pairs" = integer(), "Year*" = integer())
)

analysis <- analyse(model, data = bauw::peregrine)

coef(analysis)
```

```{r, message = FALSE}
year <- predict(analysis, new_data = "Year")

ggplot(data = year, aes(x = Year, y = estimate)) +
  geom_point(data = bauw::peregrine, aes(y = Pairs)) +
  geom_line() +
  geom_line(aes(y = lower), linetype = "dotted") +
  geom_line(aes(y = upper), linetype = "dotted") +
  expand_limits(y = 0)
```

## Installation

To install from GitHub
```
# install.packages("devtools")
devtools::install_github("poissonconsulting/mbr")
```

## Contribution

Please report any [issues](https://github.com/poissonconsulting/mbr/issues).

[Pull requests](https://github.com/poissonconsulting/mbr/pulls) are always welcome.

Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.
