---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "tools/README-"
)
```

 [![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
 [![Travis-CI Build Status](https://travis-ci.org/poissonconsulting/mbr.svg?branch=master)](https://travis-ci.org/poissonconsulting/mbr)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/poissonconsulting/mbr?branch=master&svg=true)](https://ci.appveyor.com/project/poissonconsulting/mbr)
[![Coverage Status](https://img.shields.io/codecov/c/github/poissonconsulting/mbr/master.svg)](https://codecov.io/github/poissonconsulting/mbr?branch=master)
[![License: MIT](https://img.shields.io/badge/License-MIT-green.svg)](https://opensource.org/licenses/MIT)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.1162349.svg)](https://doi.org/10.5281/zenodo.1162349)

# mbr

## Introduction

`mbr` (pronounced MBR) is an R package of utility functions and virtual classes for the model builder family of R packages.

Currently there are packages which implement the functionality for JAGS ([jmbr](https://github.com/poissonconsulting/jmbr)), STAN ([smbr](https://github.com/poissonconsulting/smbr)) and TMB ([tmbr](https://github.com/poissonconsulting/tmbr)).

## Installation

To install from GitHub
```
# install.packages("devtools")
devtools::install_github("poissonconsulting/mbr")
```

## Citation

```{r, comment="", echo=FALSE}
citation(package = 'mbr')
```

## Contribution

Please report any [issues](https://github.com/poissonconsulting/mbr/issues).

[Pull requests](https://github.com/poissonconsulting/mbr/pulls) are always welcome.

Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

## Inspiration

[jaggernaut](https://github.com/poissonconsulting/jaggernaut).
