---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "tools/README-"
)
```

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/poissonconsulting/mbr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/poissonconsulting/mbr/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/github/poissonconsulting/mbr/branch/main/graph/badge.svg?token=RSs3t4kgR0)](https://codecov.io/github/poissonconsulting/mbr)
[![License: MIT](https://img.shields.io/badge/License-MIT-green.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

# mbr

## Introduction

`mbr` (pronounced MBR) is an R package of utility functions and virtual classes for the model builder family of R packages.

Currently there are packages which implement the functionality for JAGS ([jmbr](https://github.com/poissonconsulting/jmbr)), STAN ([smbr](https://github.com/poissonconsulting/smbr)) and TMB ([tmbr](https://github.com/poissonconsulting/tmbr)).

## Installation

To install the latest development version from [GitHub](https://github.com/poissonconsulting/mbr)
```r
# install.packages("remotes")
remotes::install_github("poissonconsulting/mbr")
```

## Citation

```{r, comment="", echo=FALSE}
citation(package = 'mbr')
```

## Contribution

Please report any [issues](https://github.com/poissonconsulting/mbr/issues).

[Pull requests](https://github.com/poissonconsulting/mbr/pulls) are always welcome.

## Code of Conduct

Please note that the mbr project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.


## Inspiration

[jaggernaut](https://github.com/poissonconsulting/jaggernaut)
