---
title: "Prediction Patterns with predict()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prediction Patterns with predict()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  cache.lazy = FALSE,
  fig.width = 7,
  fig.height = 4
)
set.seed(42)
```

`predict.mb_analysis()` summarises the posterior predictive distribution of a
fitted Bayesian model at new covariate values, returning a tidy data frame
with point estimates and compatibility limits. This vignette demonstrates the
core prediction patterns — covariate grids, term selection, effect sizes,
random-effect zeroing — and the lower-level tools `mcmc_derive()` and
`mcmc_derive_data()` for derived quantities and group-level summaries.

```{r libraries, message = FALSE}
library(embr)
library(smbr2)
library(newdata)
library(dplyr)
library(ggplot2)
```

## Plot helpers

Two small helpers keep plots consistent across the vignette. `pointrange` is
the default for discrete predictions (per factor level); `ribbon` is for
predictions over a continuous covariate.

```{r plot-helpers}
plot_pointrange <- function(df, x) {
  ggplot(df, aes({{ x }}, estimate, ymin = lower, ymax = upper)) +
    geom_pointrange() +
    theme_minimal()
}

plot_ribbon <- function(df, x) {
  ggplot(df, aes({{ x }}, estimate, ymin = lower, ymax = upper)) +
    geom_ribbon(alpha = 0.2) +
    geom_line() +
    theme_minimal()
}
```

## Setup: fit the shared model

The same fish-count model used in `vignette("analyse")`: 5 sites × 5 years ×
2 treatments, with an unbalanced design.

```{r simulate}
sites <- letters[1:5]
years <- 1:5
treatments <- c("control", "restored")

design <- expand.grid(
  site = sites, annual = years, treatment = treatments,
  rep = 1:3, stringsAsFactors = FALSE
) |>
  filter(!(site == "e" & annual %in% c(1, 2, 3))) |>
  mutate(
    site = factor(site, levels = sites),
    annual = factor(annual, levels = years),
    treatment = factor(treatment, levels = treatments),
    temperature = rnorm(n(), mean = 12, sd = 2)
  )

re_site <- setNames(rnorm(length(sites), 0, 0.4), sites)
re_site_annual <- matrix(
  rnorm(length(sites) * length(years), 0, 0.2),
  nrow = length(sites), dimnames = list(sites, years)
)

data <- design |>
  mutate(
    temp_c = (temperature - mean(temperature)) / sd(temperature),
    log_mu = log(40) + 0.5 * (treatment == "restored") + 0.15 * temp_c +
      re_site[as.character(site)] +
      mapply(function(s, a) re_site_annual[s, a], as.character(site), as.character(annual)),
    count = rnbinom(n(), mu = exp(log_mu), size = 5)
  ) |>
  select(count, site, annual, treatment, temperature)
```

```{r stan-code}
count_code <- "
data {
  int<lower=2> nObs;
  int<lower=1> nsite;
  int<lower=1> nannual;
  int<lower=1> ntreatment;
  array[nObs] int<lower=1> site;
  array[nObs] int<lower=1> annual;
  array[nObs] int<lower=1> treatment;
  vector[nObs] temperature;
  array[nObs] int<lower=0> count;
}
parameters {
  real bIntercept;
  vector[ntreatment - 1] bTreatment_dev;
  real bTemp;
  real<lower=0> sSite;
  real<lower=0> sSiteAnnual;
  real<lower=0> bPhi;
  vector[nsite] z_bSite;
  matrix[nsite, nannual] z_bSiteAnnual;
}
model {
  vector[nsite] bSite = z_bSite * sSite;
  matrix[nsite, nannual] bSiteAnnual = z_bSiteAnnual * sSiteAnnual;
  vector[ntreatment] bTreatment;
  bTreatment[1] = 0;
  for (k in 2:ntreatment) bTreatment[k] = bTreatment_dev[k - 1];

  vector[nObs] log_eCount;
  for (i in 1:nObs) {
    log_eCount[i] = bIntercept + bTreatment[treatment[i]] +
                    bTemp * temperature[i] + bSite[site[i]] +
                    bSiteAnnual[site[i], annual[i]];
  }

  bIntercept ~ normal(0, 5);
  bTreatment_dev ~ normal(0, 5);
  bTemp ~ normal(0, 5);
  sSite ~ exponential(1);
  sSiteAnnual ~ exponential(1);
  bPhi ~ gamma(2, 0.5);
  to_vector(z_bSite) ~ std_normal();
  to_vector(z_bSiteAnnual) ~ std_normal();

  count ~ neg_binomial_2_log(log_eCount, bPhi);
}
"

count_model <- model(
  code = count_code,
  new_expr = {
    bSite <- z_bSite * sSite
    bSiteAnnual <- z_bSiteAnnual * sSiteAnnual
    bTreatment <- c(0, bTreatment_dev)
    eBaseCount <- exp(bIntercept)
    eRestoredEffect <- exp(bTreatment_dev[1])
    for (i in 1:nObs) {
      log(eCount[i]) <- bIntercept +
        bTreatment[treatment[i]] +
        bTemp * temperature[i] +
        bSite[site[i]] +
        bSiteAnnual[site[i], annual[i]]
      prediction[i] <- eCount[i]
      fit[i] <- log(eCount[i])
      log_lik[i] <- log_lik_neg_binom(count[i], eCount[i], bPhi)
    }
  },
  new_expr_vec = TRUE,
  select_data = list(
    count = 1L,
    site = factor("a"),
    annual = factor("a"),
    treatment = factor("a"),
    `temperature*` = 1
  ),
  random_effects = list(
    z_bSite = "site",
    z_bSiteAnnual = c("site", "annual")
  )
)
```

```{r fit, message = FALSE}
analysis <- analyse(
  count_model,
  data = data,
  stan_engine = "cmdstan-mcmc",
  nchains = 3L,
  niters = 500L,
  niters_warmup = 500L,
  seed = 42L,
  parallel = TRUE,
  quiet = TRUE,
  beep = FALSE
)

data <- data_set(analysis)  # carries the scaled temperature column

coef(analysis, include_constant = FALSE, simplify = TRUE) |>
  mutate(across(estimate:svalue, ~ signif(.x, 3)))
```

## Default prediction

With no `new_data`, `predict()` returns fitted values for every row of the
analysis dataset:

```{r default-predict}
predict(analysis) |> head()
```

## Covariate grids with the `newdata` helpers

Covariates not specified in `xnew_data()` are held at a reference value:
mean for continuous, first level for factors.

### Continuous covariate

```{r grid-temperature}
pred_temp <- xnew_data(data, temperature) |>
  predict(analysis, new_data = _)

head(pred_temp)

plot_ribbon(pred_temp, temperature) +
  labs(y = "Expected count", x = "Temperature (scaled)")
```

### Specific values via `xnew_seq()`

```{r grid-seq}
xnew_data(data, xnew_seq(temperature, length_out = 5)) |>
  predict(analysis, new_data = _)
```

### Single factor

```{r grid-site}
pred_site <- xnew_data(data, site) |>
  predict(analysis, new_data = _)

pred_site

plot_pointrange(pred_site, site) +
  labs(y = "Expected count", x = "Site")
```

### Multiple factors with the hyperprior filling in

When `site` and `annual` are crossed, the full 25-row grid is returned —
even though only 22 site×annual cells were observed. Predictions for the
unobserved combinations come from the `z_bSiteAnnual` hyperprior.

```{r grid-site-annual}
xnew_data(data, site, annual) |>
  predict(analysis, new_data = _) |>
  nrow()
```

### `xobs_only()` for observed combinations only

`xobs_only()` restricts the grid to combinations actually present in the
data:

```{r grid-xobs-only}
xnew_data(data, xobs_only(site, annual)) |>
  predict(analysis, new_data = _) |>
  nrow()
```

22 rows instead of 25 — exactly the unbalanced combinations from the
design.

### Mixing factor and continuous covariates

```{r grid-mix}
xnew_data(data, treatment, temperature = 1.5) |>
  predict(analysis, new_data = _)
```

### Casting specific factor levels with `xcast()`

```{r grid-xcast}
xnew_data(
  data,
  xcast(site = "a", treatment = "restored"),
  temperature = 1
) |>
  predict(analysis, new_data = _)
```

## Term selection

`term` chooses which quantity in `new_expr` to summarise. The default is
`"prediction"`; `"fit"` returns the link-scale (log) prediction:

```{r term-fit}
xnew_data(data, treatment) |>
  predict(analysis, new_data = _, term = "fit") |>
  mutate(across(estimate:svalue, ~ signif(.x, 3)))
```

`exp(fit) = prediction` at every row, as expected from the model.

### Scalar derived quantities

Pass `new_data = character(0)` to extract a scalar quantity from `new_expr`:

```{r term-scalar}
predict(analysis, new_data = character(0), term = "eBaseCount")
predict(analysis, new_data = character(0), term = "eRestoredEffect")
```

`eBaseCount` is the expected count at reference temperature and the control
treatment (population-level). `eRestoredEffect` is the multiplicative effect
of the restoration treatment.

## Overriding `new_expr` and injecting constants

`new_expr` can be replaced inline as a string. `new_values` supplies named
scalar constants into that expression's environment. A realistic use:
back-calculate true abundance from observed counts using a known detection
probability:

```{r new-expr-override}
xnew_data(data, site) |>
  predict(analysis, new_data = _) |>
  head(2)

predict(
  analysis,
  new_data = xnew_data(data, site),
  new_expr = "
    bSite <- z_bSite * sSite
    for (i in 1:length(site)) {
      prediction[i] <- exp(bIntercept + bSite[site[i]]) / p_detection
    }
  ",
  new_values = list(p_detection = 0.6)
) |>
  head(2)
```

The second call divides the predicted observed count by a detection
probability of 0.6, giving the estimated true abundance. Compatibility
limits widen proportionally.

## Reference data and effect sizes

When `ref_data` is a 1-row data frame, `predict()` computes
`ref_fun2(c(ref_iter, new_iter))` per MCMC iteration and summarises the
result. The returned summary describes the posterior of the *transformed*
quantity, not of the raw prediction. The default `ref_fun2` is
`proportional_change2` from the `extras` package (`(new - ref) / ref`).

### Proportional change vs control

```{r ref-prop}
ref_control <- xnew_data(data, xcast(treatment = "control"))

pred_prop <- xnew_data(data, treatment) |>
  predict(analysis, new_data = _, ref_data = ref_control)

pred_prop

plot_pointrange(pred_prop, treatment) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Change vs control", x = "Treatment")
```

The control row sits at exactly 0% by construction. The restored row gives
the proportional change in expected count.

### Absolute difference

```{r ref-diff}
xnew_data(data, treatment) |>
  predict(
    analysis,
    new_data = _,
    ref_data = ref_control,
    ref_fun2 = function(x) x[2] - x[1]
  )
```

### Log ratio

```{r ref-logratio}
xnew_data(data, treatment) |>
  predict(
    analysis,
    new_data = _,
    ref_data = ref_control,
    ref_fun2 = function(x) log(x[2] / x[1])
  )
```

Log ratios are symmetric (a 50% increase and a 33% decrease have the same
magnitude on the log scale) which makes them well-behaved when averaging
effects across studies; proportional change is bounded below at -100% but
unbounded above.

## Random-effect zeroing

A `z_b*` parameter is zeroed (collapsing predictions to the population
level) when *any* of its associated factor columns in `new_data` is entirely
at level 1. `random_effects` controls which `z_b*` parameters are candidates
for zeroing.

### `NULL` (default), site omitted

`xnew_data(data, temperature)` leaves both `site` and `annual` at level 1.
Both `z_bSite` and `z_bSiteAnnual` are zeroed → population-level prediction:

```{r re-null-site-omitted}
xnew_data(data, temperature) |>
  predict(analysis, new_data = _) |>
  head(3)
```

### `NULL`, site varied

`xnew_data(data, site)` varies `site` but `annual` is still at level 1.
`z_bSite` is kept (site varies), but `z_bSiteAnnual` is still zeroed because
`annual` triggers the any-factor rule:

```{r re-null-site-varied}
xnew_data(data, site) |>
  predict(analysis, new_data = _, random_effects = NULL)
```

### `FALSE`: skip all zeroing

```{r re-false}
xnew_data(data, site) |>
  predict(analysis, new_data = _, random_effects = FALSE)
```

Both random effects are kept regardless of the factor levels. The point
estimates shift slightly relative to the `NULL` case because the
`z_bSiteAnnual` posterior centred at zero is now propagated rather than
zeroed.

### Named list: zero only one parameter

To zero only `z_bSite` (keep `z_bSiteAnnual` regardless of factor levels):

```{r re-named-site}
xnew_data(data, site) |>
  predict(
    analysis,
    new_data = _,
    random_effects = list(z_bSite = "site")
  )
```

The opposite — zero only `z_bSiteAnnual` (`z_bSite` always kept):

```{r re-named-site-annual}
xnew_data(data, site) |>
  predict(
    analysis,
    new_data = _,
    random_effects = list(z_bSiteAnnual = c("site", "annual"))
  )
```

The rule across all four cases: a multi-factor random effect is zeroed when
*any* (OR, not AND) of its factor columns is constant at level 1; the named
list narrows the set of candidates considered.

## Scalar derived quantities with `mcmc_derive()`

`mcmc_derive()` returns the raw MCMC samples for one or more scalar terms in
`new_expr`. Useful when you want to compute functions of parameters with
full uncertainty propagation.

### Pull related scalars by regex

Both `eBaseCount` and `eRestoredEffect` are extracted in one call, sharing
the same iteration ordering:

```{r mcmc-derive-regex}
scalars <- mcmc_derive(
  analysis,
  new_data = character(0),
  term = "^(eBaseCount|eRestoredEffect)$"
)

coef(scalars)
```

### Arithmetic on `mcmcr` objects propagates uncertainty

Multiplying `mcmcr` objects acts element-wise across MCMC iterations:

```{r mcmc-derive-arithmetic}
eExpectedRestored <- scalars$eBaseCount * scalars$eRestoredEffect
coef(eExpectedRestored)
```

This is the absolute expected count at reference site and temperature under
the restored treatment — the analytic posterior of a function of parameters,
not a posterior over a per-row prediction grid.

### Custom expression + injected constants

A common use is converting counts to biomass with an externally measured
mean fish mass. No refit required:

```{r mcmc-derive-custom}
as.mcmcr(analysis) |>
  mcmc_derive(
    expr = "biomass <- exp(bIntercept) * mean_mass_g",
    values = list(mean_mass_g = 12),
    silent = TRUE
  ) |>
  coef()
```

## Group-level summaries with `mcmc_derive_data()`

`mcmc_derive_data()` pairs MCMC samples with the per-row data and returns an
`mcmc_data` object. `group_by()` + `summarise()` then aggregates across rows
within each group per MCMC iteration — propagating posterior correlations
that ad-hoc summarisation of point estimates would discard.

### Total expected count per treatment

```{r mcd-total}
mcmc_derive_data(analysis, new_data = data, term = "^eCount$") |>
  group_by(treatment) |>
  summarise(.fun = sum) |>
  coef()
```

The default `.fun = sum` returns the posterior of *total* abundance per
treatment across all rows. Summing point estimates after `predict()` would
give the same central tendency but understated uncertainty.

### Per-site mean abundance

```{r mcd-per-site}
mcmc_derive_data(analysis, new_data = data, term = "^eCount$") |>
  group_by(site) |>
  summarise(.fun = mean) |>
  coef()
```

### Custom `.fun`: between-site CV per treatment

A coefficient of variation across sites measures spatial heterogeneity in
expected counts. Computed within each MCMC iteration to propagate
uncertainty into the variability metric:

```{r mcd-custom-fun}
mcmc_derive_data(analysis, new_data = data, term = "^eCount$") |>
  group_by(treatment) |>
  summarise(.fun = function(x) sd(x) / mean(x)) |>
  coef()
```

## Combining across analyses with `combine_samples()`

`mcmcr::combine_samples()` (with the `mcmc_data` method from `mcmcdata`)
joins two `mcmc_data` objects on shared data columns and applies a function
to align MCMC draws row-by-row. Its natural use is composing quantities
derived from *independent* analyses: for example, multiplying an abundance
posterior by a mass-per-fish posterior to get biomass, where each component
was fitted from a different dataset.

A sketch — assume a separate `mass_analysis` was fitted to size-measurement
data (different observation model, same site factor), and both analyses
expose a per-site quantity:

```{r combine-samples, eval = FALSE}
# Abundance posterior per site, from the counts model fitted above
abundance <- mcmc_derive_data(
  analysis,
  new_data = xnew_data(data, site),
  term = "^eCount$"
)

# Mass-per-fish posterior per site, from a separate analysis (sketch)
mass_per_fish <- mcmc_derive_data(
  mass_analysis,
  new_data = xnew_data(mass_data, site),
  term = "^eMass$"
)

# Per-site biomass posterior: abundance * mass-per-fish
biomass <- mcmcr::combine_samples(
  abundance,
  mass_per_fish,
  by = "site",
  fun = prod
)

coef(biomass)
```

`by` names the data columns to join on (here `site`); rows present in both
inputs are paired. `fun` is applied to the matched MCMC draws — `prod` gives
the row-wise product across iterations. The result is an `mcmc_data` with
one row per joined key and a propagated posterior in `biomass`.

Within a single analysis, prefer expressing the composition directly in
`new_expr` — `combine_samples()` earns its keep when the components come
from independently fitted models whose draws need joining without re-running
them.

## Cheat sheet

| Goal | Tool |
|---|---|
| Posterior summary at new covariate values | `predict()` |
| Compatibility intervals for a derived quantity | `predict()` with `new_expr` override |
| Effect size (proportional, absolute, log ratio) | `predict()` with `ref_data` + `ref_fun2` |
| Scalar parameter posteriors and arithmetic | `mcmc_derive()` |
| Group-level aggregates (sum, mean, custom) per MCMC iteration | `mcmc_derive_data()` + `group_by()` + `summarise()` |
| Combine quantities derived from independent analyses | `mcmcr::combine_samples()` |

See `?predict.mb_analysis`, `?mcmc_derive.mb_analysis`,
`?mcmc_derive_data.mb_analysis`, and `?mcmcr::combine_samples` for full
argument documentation.
