---
title: "Fitting Models with analyse()"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  cache.lazy = FALSE,
  fig.width = 7,
  fig.height = 4
)
set.seed(42)
```

`analyse()` is embr's main entry point for fitting a Bayesian hierarchical
model to data using Stan (via rstan or cmdstanr) or JAGS (via rjags). This
guide walks through the primary `analyse.mb_model` method on a simple example
fish count model, then briefly covers the two convenience methods
`analyse.mb_models` (for fitting a list of models) and `analyse.character`
(for fitting from a raw code string).

```{r libraries, message = FALSE}
library(embr)
library(smbr2)
library(dplyr)
```

## Example model

The example model is a fish count model, where data has counts at 5 sites
across 5 annual periods under two treatments (`control`, `restored`), with
stream temperature as a continuous covariate. Sampling is unbalanced: 4 sites
are monitored in all 5 years; the 5th site is sampled in only 2 years.

```{r simulate}
sites <- letters[1:5]
years <- 1:5
treatments <- c("control", "restored")

design <- expand.grid(
  site = sites,
  annual = years,
  treatment = treatments,
  rep = 1:3,
  stringsAsFactors = FALSE
) |>
  filter(!(site == "e" & annual %in% c(1, 2, 3))) |>
  mutate(
    site = factor(site, levels = sites),
    annual = factor(annual, levels = years),
    treatment = factor(treatment, levels = treatments),
    temperature = rnorm(n(), mean = 12, sd = 2)
  )

bIntercept_true <- log(40)
bTreatment_true <- 0.5
bTemp_true <- 0.15
sSite_true <- 0.4
sSiteAnnual_true <- 0.2
bPhi_true <- 5

re_site <- setNames(rnorm(length(sites), 0, sSite_true), sites)
re_site_annual <- matrix(
  rnorm(length(sites) * length(years), 0, sSiteAnnual_true),
  nrow = length(sites),
  dimnames = list(sites, years)
)

data <- design |>
  mutate(
    temp_c = (temperature - mean(temperature)) / sd(temperature),
    log_mu = bIntercept_true +
      bTreatment_true * (treatment == "restored") +
      bTemp_true * temp_c +
      re_site[as.character(site)] +
      mapply(
        function(s, a) re_site_annual[s, a],
        as.character(site),
        as.character(annual)
      ),
    count = as.integer(rnbinom(n(), mu = exp(log_mu), size = bPhi_true))
  ) |>
  select(count, site, annual, treatment, temperature)

head(data)
```

## Defining the model

The model is defined in two parts: the Stan code (likelihood and priors) and
an R `new_expr` (used post-fitting for predictions, residuals, sensitivity, and the log likelihood). `select_data` declares the columns to extract from `data` and how to validate / transform them. `random_effects` maps the relevant
parameters (e.g., `z_b*` in parameterization below) to their factor columns so embr can zero them at prediction time to get predictions for 'typical' values. 

```{r stan-code}
count_code <- "
data {
  int<lower=2> nObs;
  int<lower=1> nsite;
  int<lower=1> nannual;
  int<lower=1> ntreatment;
  array[nObs] int<lower=1> site;
  array[nObs] int<lower=1> annual;
  array[nObs] int<lower=1> treatment;
  vector[nObs] temperature;
  array[nObs] int<lower=0> count;
}
parameters {
  real bIntercept;
  vector[ntreatment - 1] bTreatment_dev;
  real bTemp;
  real<lower=0> sSite;
  real<lower=0> sSiteAnnual;
  real<lower=0> bPhi;
  vector[nsite] z_bSite;
  matrix[nsite, nannual] z_bSiteAnnual;
}
model {
  vector[nsite] bSite = z_bSite * sSite;
  matrix[nsite, nannual] bSiteAnnual = z_bSiteAnnual * sSiteAnnual;
  vector[ntreatment] bTreatment;
  bTreatment[1] = 0;
  for (k in 2:ntreatment) bTreatment[k] = bTreatment_dev[k - 1];

  vector[nObs] log_eCount;
  for (i in 1:nObs) {
    log_eCount[i] = bIntercept + bTreatment[treatment[i]] +
                    bTemp * temperature[i] + bSite[site[i]] +
                    bSiteAnnual[site[i], annual[i]];
  }

  bIntercept ~ normal(0, 5);
  bTreatment_dev ~ normal(0, 5);
  bTemp ~ normal(0, 5);
  sSite ~ exponential(1);
  sSiteAnnual ~ exponential(1);
  bPhi ~ gamma(2, 0.5);
  to_vector(z_bSite) ~ std_normal();
  to_vector(z_bSiteAnnual) ~ std_normal();

  count ~ neg_binomial_2_log(log_eCount, bPhi);
}
"

count_model <- model(
  code = count_code,
  new_expr = {
    bSite <- z_bSite * sSite
    bSiteAnnual <- z_bSiteAnnual * sSiteAnnual
    bTreatment <- c(0, bTreatment_dev)
    eBaseCount <- exp(bIntercept)
    eRestoredEffect <- exp(bTreatment_dev[1])
    for (i in 1:nObs) {
      log(eCount[i]) <- bIntercept +
        bTreatment[treatment[i]] +
        bTemp * temperature[i] +
        bSite[site[i]] +
        bSiteAnnual[site[i], annual[i]]
      prediction[i] <- eCount[i]
      fit[i] <- eCount[i]
      log_lik[i] <- log_lik_neg_binom(count[i], eCount[i], bPhi)
    }
  },
  new_expr_vec = TRUE,
  select_data = list(
    count = 1L,
    site = factor("a"),
    annual = factor("a"),
    treatment = factor("a"),
    `temperature*` = 1
  ),
  random_effects = list(
    z_bSite = "site",
    z_bSiteAnnual = c("site", "annual")
  )
)
```

A few notes on the `model()` arguments:

- `code`: the Stan model code.
- `new_expr`: an R expression evaluated post-fitting against MCMC draws. It
  defines `prediction[i]` (used by `predict()`), `fit[i]` (used by
  `fitted()` and for residuals), and `log_lik[i]` (used by IC and
  prior-sensitivity tools). The scale of `fit[i]` is model-specific; define
  it to match what your `res_*` and `log_lik_*` helpers expect. Scalar
  derived quantities such as `eBaseCount` and `eRestoredEffect` can be
  extracted directly with `predict(analysis, new_data = character(0),
  term = "eBaseCount")`.
- `select_data`: the columns required from `data`. A trailing `*` on
  `temperature` requests scaling; the original `temperature`
  column is replaced with the scaled version in the analysis dataset.
- `random_effects`: tells embr which factors each non-centred random-effect
  parameter depends on. `predict()` uses this to decide when to zero `z_b*`
  parameters (see the prediction vignette).

## Fitting with `analyse.mb_model`

The default backend is rstan for backwards-compatibility. However, it is recommended for Stan to use cmdstanr's MCMC sampler via smbr2 package by passing
`stan_engine = "cmdstan-mcmc"`. We also set `niters`, `niters_warmup`, and `nchains` although this is also typically handled by the `set_analysis_mode()` function. 

```{r fit-default}
analysis <- analyse(
  count_model,
  data = data,
  stan_engine = "cmdstan-mcmc",
  nchains = 3L,
  niters = 500L,
  niters_warmup = 500L,
  seed = 42L,
  parallel = TRUE,
  quiet = TRUE,
  beep = FALSE
)

glance(analysis)
```

`coef()` extracts the fixed-effect and scale parameters:

```{r coef}
coef(analysis, include_constant = FALSE, simplify = TRUE) |>
  mutate(across(estimate:svalue, ~ signif(.x, 3)))
```

## Common arguments

The following arguments cover most day-to-day use of `analyse.mb_model`:

| Argument | Purpose |
|---|---|
| `nchains` | Number of MCMC chains. Default 3. |
| `niters` | Number of sampling iterations saved per chain. |
| `niters_warmup` | Number of warmup iterations (Stan only). Defaults to `niters`. |
| `nthin` | Thinning interval. Defaults to the value set in `model()`. |
| `seed` | Positive integer for reproducible sampling. |
| `parallel` | Run chains in parallel (`foreach` backend). |
| `quiet` | Suppress sampling progress messages. |
| `glance` | Print a one-line model summary after fitting. |
| `beep` | Beep on completion (helpful for long fits). |

Stan-only options (`niters_warmup`, `seed`) are ignored by JAGS models.

## Engine dispatch

The `stan_engine` argument selects the Stan backend for Stan models. JAGS
models always go through `rjags` regardless of this argument.

| `stan_engine` | Backend | Use case |
|---|---|---|
| `character(0)` (default) | `rstan::sampling()` | Default MCMC; rstan-only options like `control = list(adapt_delta = ...)`. |
| `"cmdstan-mcmc"` | `cmdstanr::sample()` | Full MCMC via cmdstanr. Faster compilation, better diagnostics than rstan. |
| `"cmdstan-pathfinder"` | `cmdstanr::pathfinder()` | Fast variational approximation; useful for initialisation or rough exploration. |
| `"cmdstan-variational"` | `cmdstanr::variational()` | ADVI; faster than MCMC, less accurate for hierarchical models. |
| `"cmdstan-optimize"` | `cmdstanr::optimize()` | Point estimate (MAP); no posterior uncertainty. |
| `"cmdstan-laplace"` | `cmdstanr::laplace()` | Laplace approximation around the mode. |

For each cmdstanr engine, embr sets a fixed set of cmdstanr arguments
(`chains`, `iter_warmup`, `iter_sampling`, `data`, `seed`, etc.) from its
own parameters. Other cmdstanr arguments can be passed through `...`. See
`?analyse` for the exact list of embr-controlled arguments per engine.

A pathfinder fit on the same model takes a fraction of the time, but posterior distributions will be less reliable, especially for complex hierarchical models:

```{r fit-pathfinder}
analysis_pf <- analyse(
  count_model,
  data = data,
  stan_engine = "cmdstan-pathfinder",
  niters = 500L,
  seed = 42L,
  quiet = TRUE,
  beep = FALSE
)

coef(analysis_pf, include_constant = FALSE, simplify = TRUE) |>
  filter(term %in% c("bIntercept", "bTreatment_dev[1]", "bTemp")) |>
  mutate(across(estimate:svalue, ~ signif(.x, 3)))
```

## Engine-specific arguments via `...`

Additional cmdstanr arguments are passed through `...`. For example, to
tighten the NUTS step size:

```{r fit-adapt, eval = FALSE}
analyse(
  count_model,
  data = data,
  stan_engine = "cmdstan-mcmc",
  adapt_delta = 0.95,
  max_treedepth = 12,
  seed = 42L
)
```

`init` is the one embr-supplied argument that *can* be overridden via `...`.
Supply a zero-argument function returning a named list of initial values per
chain:

```{r fit-init, eval = FALSE}
init_fun <- function() list(bIntercept = log(40))

analyse(
  count_model,
  data = data,
  stan_engine = "cmdstan-mcmc",
  init = init_fun,
  seed = 42L
)
```

## In brief: `analyse(mb_models, ...)`

To fit several model structures against the same data, wrap them in
`models()` and call `analyse()` once:

```{r mb-models, eval = FALSE}
count_model_no_temp <- model(
  code = count_code, # in practice, with bTemp dropped
  new_expr = {
    # ... no temperature term ...
  },
  select_data = list(
    count = 1L,
    site = factor("a"),
    annual = factor("a"),
    treatment = factor("a")
  ),
  random_effects = list(
    z_bSite = "site",
    z_bSiteAnnual = c("site", "annual")
  )
)

candidate_models <- models(
  with_temp = count_model,
  without_temp = count_model_no_temp
)

analyses <- analyse(
  candidate_models,
  data = data,
  stan_engine = "cmdstan-mcmc",
  seed = 42L
)
```

The result is an `mb_analyses` list. `predict.mb_analyses()` returns
IC-weighted model-averaged predictions across the list.

## In brief: `analyse(character, ...)`

For quick prototyping without a separate `model()` object, pass a Stan or
JAGS code string directly to `analyse()` with a `select_data` list:

```{r mb-character, eval = FALSE}
analyse(
  count_code,
  data = data,
  select_data = list(
    count = 1L,
    site = factor("a"),
    annual = factor("a"),
    treatment = factor("a"),
    `temperature*` = 1
  ),
  stan_engine = "cmdstan-mcmc",
  seed = 42L
)
```

This is a convenience wrapper: it calls `model()` internally and dispatches
to the `mb_model` method. Use it for one-shot fits; use the explicit
`model()` form when you want to reuse, modify, or inspect the model object.

## Next steps

- See `?analyse` for the full per-engine argument documentation.
- See the [prediction article](prediction.html) for posterior predictive
  summaries, effect-size calculations, and the `mcmc_derive` /
  `mcmc_derive_data` tools for derived quantities.
- `coef()`, `glance()`, `residuals()`, and `plot_residuals()` are the
  standard inspection tools for an `mb_analysis`.
