---
title: "Fitting Models with analyse()"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  cache.lazy = FALSE,
  fig.width = 7,
  fig.height = 4
)
set.seed(42)
```

`analyse()` is used to fit a Bayesian hierarchical
model to data using Stan (via rstan or cmdstanr) or JAGS (via rjags). This
guide walks through the primary `analyse.mb_model` method on an example model, then briefly covers the two convenience methods
`analyse.mb_models` (for fitting a list of models) and `analyse.character`
(for fitting from a raw code string).

```{r libraries, message = FALSE}
library(embr)
library(smbr2)
library(dplyr)
```

## Example model

The example is a fish count model, where data has counts at 5 sites
across 5 annual periods under two treatments (`control`, `restored`), with
stream temperature as a continuous covariate. Sampling is unbalanced: 4 sites
are monitored in all 5 years; the 5th site is sampled in only 2 years.

```{r simulate}
sites <- letters[1:5]
years <- 1:5
treatments <- c("control", "restored")

design <- expand.grid(
  site = sites,
  annual = years,
  treatment = treatments,
  rep = 1:3,
  stringsAsFactors = FALSE
) |>
  filter(!(site == "e" & annual %in% c(1, 2, 3))) |>
  mutate(
    site = factor(site, levels = sites),
    annual = factor(annual, levels = years),
    treatment = factor(treatment, levels = treatments),
    temperature = rnorm(n(), mean = 12, sd = 2)
  )

bIntercept_true <- log(40)
bTreatment_true <- 0.5
bTemp_true <- 0.15
sSite_true <- 0.4
sSiteAnnual_true <- 0.2
bPhi_true <- 5

re_site <- setNames(rnorm(length(sites), 0, sSite_true), sites)
re_site_annual <- matrix(
  rnorm(length(sites) * length(years), 0, sSiteAnnual_true),
  nrow = length(sites),
  dimnames = list(sites, years)
)

data <- design |>
  mutate(
    temp_c = (temperature - mean(temperature)) / sd(temperature),
    log_mu = bIntercept_true +
      bTreatment_true * (treatment == "restored") +
      bTemp_true * temp_c +
      re_site[as.character(site)] +
      mapply(
        function(s, a) re_site_annual[s, a],
        as.character(site),
        as.character(annual)
      ),
    count = as.integer(ran_neg_binom(n(), mu = exp(log_mu), size = bPhi_true))
  ) |>
  select(count, site, annual, treatment, temperature)

head(data)
```

## Defining the model

The model is defined in two parts: the Stan code (likelihood and priors) and
an R `new_expr` (used post-fitting for predictions, residuals, sensitivity, and log likelihood). `select_data` declares the columns to extract from `data` and how to validate / transform them. `random_effects` maps the relevant
parameters (e.g., `z_b*` in parameterization below) to their factor columns so embr can zero them at prediction time to get predictions for a 'typical' group level.

```{r stan-code}
count_code <- "
data {
  int<lower=2> nObs;
  int<lower=1> nsite;
  int<lower=1> nannual;
  int<lower=1> ntreatment;
  array[nObs] int<lower=1> site;
  array[nObs] int<lower=1> annual;
  array[nObs] int<lower=1> treatment;
  vector[nObs] temperature;
  array[nObs] int<lower=0> count;
}
parameters {
  real bIntercept;
  vector[ntreatment - 1] bTreatment_dev;
  real bTemp;
  real<lower=0> sSite;
  real<lower=0> sSiteAnnual;
  real<lower=0> bPhi;
  vector[nsite] z_bSite;
  matrix[nsite, nannual] z_bSiteAnnual;
}
model {
  vector[nsite] bSite = z_bSite * sSite;
  matrix[nsite, nannual] bSiteAnnual = z_bSiteAnnual * sSiteAnnual;
  vector[ntreatment] bTreatment;
  bTreatment[1] = 0;
  for (k in 2:ntreatment) bTreatment[k] = bTreatment_dev[k - 1];

  vector[nObs] log_eCount;
  for (i in 1:nObs) {
    log_eCount[i] = bIntercept + bTreatment[treatment[i]] +
                    bTemp * temperature[i] + bSite[site[i]] +
                    bSiteAnnual[site[i], annual[i]];
  }

  bIntercept ~ normal(0, 5);
  bTreatment_dev ~ normal(0, 5);
  bTemp ~ normal(0, 5);
  sSite ~ exponential(1);
  sSiteAnnual ~ exponential(1);
  bPhi ~ gamma(2, 0.5);
  to_vector(z_bSite) ~ std_normal();
  to_vector(z_bSiteAnnual) ~ std_normal();

  count ~ neg_binomial_2_log(log_eCount, bPhi);
}
"

count_model <- model(
  code = count_code,
  new_expr = {
    bSite <- z_bSite * sSite
    bSiteAnnual <- z_bSiteAnnual * sSiteAnnual
    bTreatment <- c(0, bTreatment_dev)
    eBaseCount <- exp(bIntercept)
    eRestoredEffect <- exp(bTreatment_dev[1])
    for (i in 1:nObs) {
      log(eCount[i]) <- bIntercept +
        bTreatment[treatment[i]] +
        bTemp * temperature[i] +
        bSite[site[i]] +
        bSiteAnnual[site[i], annual[i]]
      prediction[i] <- eCount[i]
      fit[i] <- eCount[i]
      log_lik[i] <- log_lik_neg_binom(count[i], eCount[i], bPhi)
    }
    lprior[1] <- log_lik_norm(bIntercept, 0, 5)
    lprior[2] <- sum(log_lik_norm(bTreatment_dev, 0, 5))
    lprior[3] <- log_lik_norm(bTemp, 0, 5)
    lprior[4] <- log_lik_exp(sSite, 1)
    lprior[5] <- log_lik_exp(sSiteAnnual, 1)
    lprior[6] <- log_lik_gamma(bPhi, 2, 0.5)
  },
  new_expr_vec = TRUE,
  select_data = list(
    count = 1L,
    site = factor("a"),
    annual = factor("a"),
    treatment = factor("a"),
    `temperature*` = c(0, 100)
  ),
  random_effects = list(
    z_bSite = "site",
    z_bSiteAnnual = c("site", "annual")
  )
)
```

A few notes on the `model()` arguments:

- `code`: the Stan model code.
- `new_expr`: an R expression evaluated post-fitting against MCMC draws. It
  defines `prediction[i]` (used by `predict()`), `fit[i]` (used by
  `fitted()` and for residuals), and `log_lik[i]` (used by model comparison and
  prior-sensitivity tools). The scale of `fit[i]` is model-specific; define
  it to match what your `res_*` and `log_lik_*` helpers expect. Scalar
  derived quantities such as `eBaseCount` and `eRestoredEffect` can be
  extracted directly with `predict(analysis, new_data = character(0),
  term = "eBaseCount")`.
- `select_data`: the columns required from `data`, validation checks on expected type or range of values, and rescaling transformations. 

Named list mapping data column names to type or range specifications: `1L` for integer, `1`
#' for numeric, `factor("")` for factor, `TRUE` for logical. Where an
#' explicit check is wanted, use a range form such as `c(0L, 100L)` and
#' append `NA` to allow missing values: `c(0L, 100L, NA)`.

A trailing suffix on a
  name requests a transformation; the original column is replaced with the
  transformed version in the analysis dataset. Available suffixes:

  | Suffix | Transformation | Typical use |
  |---|---|---|
  | (none) | Raw, untransformed | Response variables, factors, logicals, raw data for in-model transforms |
  | `+` | Subtract the mean (center) | Year or date variables, interpretable intercept at mean |
  | `-` | Subtract the minimum (shift to 0) | Year or date, intercept at first year |
  | `=` | Subtract the minimum and add 1 (shift to 1) | Indices that must start at 1 |
  | `/` | Divide by SD (scale) | When preserving the offset matters but units differ |
  | `*` | Subtract mean and divide by SD (standardize) | Continuous covariates; improves convergence, makes prior specification simpler and makes coefficients comparable |

  `*` is the default choice for continuous covariates. 
- `random_effects`: tells embr which factors each random-effect parameter
  is indexed by. `predict()` uses this to decide when to zero
  parameters (see the prediction article).

## Fitting with `analyse.mb_model`

rstan is the default fallback when no `stan_engine` is set; for new projects,
`stan_engine = "cmdstan-mcmc"` via smbr2 is recommended for faster compilation
and better diagnostics. `niters`, `niters_warmup`, and `nchains` are typically
set via `set_analysis_mode()`; here they are specified within the function to keep the
example self-contained.

```{r fit-default}
analysis <- analyse(
  count_model,
  data = data,
  stan_engine = "cmdstan-mcmc",
  nchains = 3L,
  niters = 500L,
  seed = 42L,
  parallel = TRUE,
  quiet = TRUE,
  beep = FALSE
)

glance(analysis)
```

`coef()` extracts the fixed-effect and scale parameters:

```{r coef}
coef(analysis, include_constant = FALSE, simplify = TRUE) |>
  mutate(across(estimate:svalue, ~ signif(.x, 3)))
```

## Common arguments

The following arguments cover most typical use of `analyse.mb_model`:

| Argument | Purpose |
|---|---|
| `nchains` | Number of MCMC chains. Default 3. |
| `niters` | Number of sampling iterations saved per chain. |
| `niters_warmup` | Number of warmup iterations (Stan only). Defaults to `niters`. |
| `nthin` | Thinning interval. Defaults to the value set in `model()`. |
| `seed` | Positive integer for reproducible sampling. |
| `parallel` | Run chains in parallel (`foreach` backend). |
| `quiet` | Suppress sampling progress messages. |
| `glance` | Print a one-line model summary after fitting. |
| `beep` | Beep on completion (helpful for long fits). |

Stan-only options (`niters_warmup`, `seed`) are ignored by JAGS models.

`niters` is the number of saved samples per chain *after* thinning, not
the total iterations run. Per chain:

- Warmup iterations (discarded): `niters_warmup`
- Post-warmup iterations: `niters * nthin`
- Saved samples: `niters` (every `nthin`-th post-warmup iteration)
- Total iterations run: `niters_warmup + niters * nthin`

Across all chains the saved posterior has `nchains * niters` draws. So
with `nchains = 3`, `niters = 500`, `niters_warmup = 500`, `nthin = 2`,
each chain runs `500 + 500 * 2 = 1500` iterations and contributes 500
saved samples, for `1500` total posterior draws.

## Analysis mode

`set_analysis_mode()` sets a bundle of session-wide defaults
(via options) so individual `analyse()` calls don't need to
respecify the same arguments. The mode controls `nchains`, `niters`,
`nthin`, `parallel`, `quiet`, `glance`, `beep`, the reanalysis loop
(`nreanalyses`, `rhat`, `esr`, `duration`), and `conf_level`. Typically set
once at the top of an analysis script.

| Mode | When to use | `nchains` | `niters` | `rhat` | `esr` |
|---|---|---|---|---|---|
| `"debug"` | Rapidly identify problems with a model definition | 2 | 10 | 1.05 | 0.1 |
| `"quick"` | Quickly test that code runs | 2 | 10 | 1.05 | 0.1 |
| `"check"` | Run when checking a package | 2 | 500 | 1.0 | 1.0 |
| `"report"` | Produce results for a report | 3 | 500 | 1.05 | 0.1 |
| `"paper"` | Produce results for a peer-reviewed paper | 4 | 1000 | 1.01 | 0.25 |
| `"reset"` | Reset all options to `NULL` (per-call defaults) | — | — | — | — |

`rhat` is the maximum acceptable R-hat (closer to 1 is stricter); `esr` is
the minimum effective sample rate (effective sample size / total samples;
closer to 1 is stricter).

Typical workflow:

```{r set-analysis-mode, eval = FALSE}
set_analysis_mode("quick") # check that model basically runs without error
analyse(count_model, data = data, stan_engine = "cmdstan-mcmc")

set_analysis_mode("report") # final fit and iterative model development
analyse(count_model, data = data, stan_engine = "cmdstan-mcmc")
```

`get_analysis_mode()` returns the current option values. An explicit
argument passed to `analyse()` always overrides the corresponding option.

## Engine dispatch

The `stan_engine` argument selects the Stan backend. JAGS models always go
through `rjags` regardless of this argument.

| `stan_engine` | Backend | Use case |
|---|---|---|
| `character(0)` (default) | `rstan::sampling()` | Default MCMC; rstan-only options like `control = list(adapt_delta = ...)`. |
| `"cmdstan-mcmc"` | `cmdstanr::sample()` | Full MCMC via cmdstanr. Faster compilation, better diagnostics than rstan. |
| `"cmdstan-pathfinder"` | `cmdstanr::pathfinder()` | Fast variational approximation; useful for initialisation or rough exploration. |
| `"cmdstan-variational"` | `cmdstanr::variational()` | ADVI; faster than MCMC, less accurate for hierarchical models. |
| `"cmdstan-optimize"` | `cmdstanr::optimize()` | Point estimate (MAP); no posterior uncertainty. |
| `"cmdstan-laplace"` | `cmdstanr::laplace()` | Laplace approximation around the mode. |

See `?analyse` for the per-engine list of embr-controlled vs pass-through
arguments.

A pathfinder fit on the same model is much faster but less reliable, especially for complex hierarchical models:

```{r fit-pathfinder}
analysis_pf <- analyse(
  count_model,
  data = data,
  stan_engine = "cmdstan-pathfinder",
  niters = 500L,
  seed = 42L,
  quiet = TRUE,
  beep = FALSE
)

coef(analysis_pf, include_constant = FALSE, simplify = TRUE) |>
  filter(term %in% c("bIntercept", "bTreatment_dev[1]", "bTemp")) |>
  mutate(across(estimate:svalue, ~ signif(.x, 3)))
```

## Engine-specific arguments via `...`

Additional cmdstanr arguments are passed through `...`. For example, to
tighten the NUTS step size:

```{r fit-adapt, eval = FALSE}
analyse(
  count_model,
  data = data,
  stan_engine = "cmdstan-mcmc",
  adapt_delta = 0.95,
  max_treedepth = 12,
  seed = 42L
)
```

`init` is the one embr-supplied argument that *can* be overridden via `...`.
Supply a zero-argument function returning a named list of initial values per
chain:

```{r fit-init, eval = FALSE}
init_fun <- function() list(bIntercept = log(40))

analyse(
  count_model,
  data = data,
  stan_engine = "cmdstan-mcmc",
  init = init_fun,
  seed = 42L
)
```

## In brief: `analyse(mb_models, ...)`

To fit several model structures against the same data, wrap them in
`models()` and call `analyse()` once:

```{r mb-models, eval = FALSE}
count_model_no_temp <- model(
  code = count_code, # in practice, with bTemp dropped
  new_expr = {
    # ... no temperature term ...
  },
  select_data = list(
    count = 1L,
    site = factor("a"),
    annual = factor("a"),
    treatment = factor("a")
  ),
  random_effects = list(
    z_bSite = "site",
    z_bSiteAnnual = c("site", "annual")
  )
)

candidate_models <- models(
  with_temp = count_model,
  without_temp = count_model_no_temp
)

analyses <- analyse(
  candidate_models,
  data = data,
  stan_engine = "cmdstan-mcmc",
  seed = 42L
)
```

The result is an `mb_analyses` list. `predict.mb_analyses()` returns
IC-weighted model-averaged predictions across the list.

## In brief: `analyse(character, ...)`

For quick prototyping without a separate `model()` object, pass a Stan or
JAGS code string directly to `analyse()` with a `select_data` list:

```{r mb-character, eval = FALSE}
analyse(
  count_code,
  data = data,
  select_data = list(
    count = 1L,
    site = factor("a"),
    annual = factor("a"),
    treatment = factor("a"),
    `temperature*` = 1
  ),
  stan_engine = "cmdstan-mcmc",
  seed = 42L
)
```

This is a convenience wrapper: it calls `model()` internally and dispatches
to the `mb_model` method. Use it for one-shot fits; use the explicit
`model()` form when you want to reuse, modify, or inspect the model object.

## Next steps

- See `?analyse` for the full per-engine argument documentation.
- See the [prediction article](prediction.html) for posterior predictive
  summaries, effect-size calculations, and the `mcmc_derive` /
  `mcmc_derive_data` tools for derived quantities.
- `coef()`, `glance()`, `residuals()`, and `plot_residuals()` are the
  standard inspection tools for an `mb_analysis`.
