---
title: "Prediction with predict() and Related Functions"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  cache.lazy = FALSE,
  fig.width = 7,
  fig.height = 4
)
set.seed(42)
```

`predict.mb_analysis()` summarizes the posterior predictive distribution of a
fitted Bayesian model at new covariate values, returning a tidy data frame
with point estimates and compatibility limits. This vignette covers:

* [Covariate grids](#covariate-grids-with-the-newdata-helpers) built with the `newdata` helpers.
* [Scalar quantities from `new_expr`](#scalar-quantities-from-new_expr).
* [Reference data and effect sizes](#reference-data-and-effect-sizes), e.g., for proportional change.
* [Predicting for an unobserved level](#predicting-for-an-unobserved-level) to include amongst-level variation.
* [Overriding `new_expr` and injecting constants](#overriding-new_expr-and-injecting-constants) for ad-hoc derived quantities.
* [Group-level summaries with `mcmc_derive_data()`](#group-level-summaries-with-mcmc_derive_data).
* [Combining across analyses with `combine_samples()`](#combining-across-analyses-with-combine_samples) for joining and summarizing samples from independent fits.
* [Scalar derived quantities with `mcmc_derive()`](#scalar-derived-quantities-with-mcmc_derive) and arithmetic on raw `mcmcr` posteriors.
* [Random-effect zeroing](#random-effect-zeroing) control.
* [Cheat sheet](#cheat-sheet) mapping common goals to tools.

```{r libraries, message = FALSE}
library(embr)
library(smbr2)
library(newdata)
library(mcmcr)
library(mcmcderive)
library(dplyr)
library(ggplot2)
```

```{r plot-helpers, include = FALSE}
plot_pointrange <- function(df, x) {
  ggplot(df, aes({{ x }}, estimate, ymin = lower, ymax = upper)) +
    geom_pointrange() +
    theme_minimal()
}

plot_ribbon <- function(df, x) {
  ggplot(df, aes({{ x }}, estimate, ymin = lower, ymax = upper)) +
    geom_ribbon(alpha = 0.2) +
    geom_line() +
    theme_minimal()
}
```

## Setup: fit demo model
A simple example fish count model is fit with Stan (via cmdstanr and smbr2). The simulated data has 5 sites × 5 years × 2 treatments, with an unbalanced design (i.e., not all sites have all years with data). 

```{r simulate, include = FALSE}
sites <- letters[1:5]
years <- 1:5
treatments <- c("control", "restored")

design <- expand.grid(
  site = sites,
  annual = years,
  treatment = treatments,
  rep = 1:3,
  stringsAsFactors = FALSE
) |>
  filter(!(site == "e" & annual %in% c(1, 2, 3))) |>
  mutate(
    site = factor(site, levels = sites),
    annual = factor(annual, levels = years),
    treatment = factor(treatment, levels = treatments),
    temperature = rnorm(n(), mean = 12, sd = 2)
  )

re_site <- setNames(rnorm(length(sites), 0, 0.4), sites)
re_site_annual <- matrix(
  rnorm(length(sites) * length(years), 0, 0.2),
  nrow = length(sites),
  dimnames = list(sites, years)
)

data <- design |>
  mutate(
    temp_c = (temperature - mean(temperature)) / sd(temperature),
    log_mu = log(40) +
      0.5 * (treatment == "restored") +
      0.15 * temp_c +
      re_site[as.character(site)] +
      mapply(
        function(s, a) re_site_annual[s, a],
        as.character(site),
        as.character(annual)
      ),
    count = as.integer(rnbinom(n(), mu = exp(log_mu), size = 5))
  ) |>
  select(count, site, annual, treatment, temperature)
```

```{r stan-code}
count_code <- "
data {
  int<lower=2> nObs;
  int<lower=1> nsite;
  int<lower=1> nannual;
  int<lower=1> ntreatment;
  array[nObs] int<lower=1> site;
  array[nObs] int<lower=1> annual;
  array[nObs] int<lower=1> treatment;
  vector[nObs] temperature;
  array[nObs] int<lower=0> count;
}
parameters {
  real bIntercept;
  vector[ntreatment - 1] bTreatment_dev;
  real bTemp;
  real<lower=0> sSite;
  real<lower=0> sSiteAnnual;
  real<lower=0> bPhi;
  vector[nsite] z_bSite;
  matrix[nsite, nannual] z_bSiteAnnual;
}
model {
  vector[nsite] bSite = z_bSite * sSite;
  matrix[nsite, nannual] bSiteAnnual = z_bSiteAnnual * sSiteAnnual;
  vector[ntreatment] bTreatment;
  bTreatment[1] = 0;
  for (k in 2:ntreatment) bTreatment[k] = bTreatment_dev[k - 1];

  vector[nObs] log_eCount;
  for (i in 1:nObs) {
    log_eCount[i] = bIntercept + bTreatment[treatment[i]] +
                    bTemp * temperature[i] + bSite[site[i]] +
                    bSiteAnnual[site[i], annual[i]];
  }

  bIntercept ~ normal(0, 5);
  bTreatment_dev ~ normal(0, 5);
  bTemp ~ normal(0, 5);
  sSite ~ exponential(1);
  sSiteAnnual ~ exponential(1);
  bPhi ~ gamma(2, 0.5);
  to_vector(z_bSite) ~ std_normal();
  to_vector(z_bSiteAnnual) ~ std_normal();

  count ~ neg_binomial_2_log(log_eCount, bPhi);
}
"

count_model <- model(
  code = count_code,
  new_expr = {
    bSite <- z_bSite * sSite
    bSiteAnnual <- z_bSiteAnnual * sSiteAnnual
    bTreatment <- c(0, bTreatment_dev)
    eBaseCount <- exp(bIntercept)
    eRestoredEffect <- exp(bTreatment_dev[1])
    for (i in 1:nObs) {
      log(eCount[i]) <- bIntercept +
        bTreatment[treatment[i]] +
        bTemp * temperature[i] +
        bSite[site[i]] +
        bSiteAnnual[site[i], annual[i]]
      prediction[i] <- eCount[i]
      fit[i] <- eCount[i]
      unobserved[i] <- exp(
        bIntercept +
          bTreatment[treatment[i]] +
          bTemp * temperature[i] +
          rnorm(1, 0, sSite) +
          rnorm(1, 0, sSiteAnnual)
      )
      log_lik[i] <- log_lik_neg_binom(count[i], eCount[i], bPhi)
    }
  },
  new_expr_vec = TRUE,
  select_data = list(
    count = 1L,
    site = factor("a"),
    annual = factor("a"),
    treatment = factor("a"),
    `temperature*` = 1
  ),
  random_effects = list(
    z_bSite = "site",
    z_bSiteAnnual = c("site", "annual")
  )
)
```

```{r fit, message = FALSE}
analysis <- analyse(
  count_model,
  data = data,
  stan_engine = "cmdstan-mcmc",
  nchains = 3L,
  niters = 500L,
  niters_warmup = 500L,
  seed = 42L,
  parallel = TRUE,
  quiet = TRUE,
  beep = FALSE
)

coef(analysis, include_constant = FALSE, simplify = TRUE) |>
  mutate(across(estimate:svalue, ~ signif(.x, 3)))
```

## Default prediction

With no `new_data`, `predict()` returns predicted values for every row of the
analysis dataset. This is pulled from the `prediction[i]` line in `new_expr`:

```{r default-predict}
predict(analysis) |> head()
```

## Covariate grids with the `newdata` helpers

Covariate grids can be generated with `newdata::xnew_data()`. Covariates not specified in `xnew_data()` are held at a reference value: mean for continuous, first level for factors.

### Continuous covariate
By default, specifying the name of a continuous covariate will generate a sequence from min to max, with length 30. Notice factors (e.g., site) are held at the first level. 

```{r grid-temperature}
temp_data <- xnew_data(data, temperature)
head(temp_data)
nrow(temp_data)

pred_temp <- predict(analysis, new_data = temp_data)

head(pred_temp)

plot_ribbon(pred_temp, temperature) +
  labs(y = "Expected count", x = "Temperature (scaled)")
```

### Specific values via `xnew_seq()` or with `=`

```{r grid-seq}
xnew_data(data, xnew_seq(temperature, length_out = 5)) |>
  predict(analysis, new_data = _)
```

```{r grid-eq}
xnew_data(data, temperature = 10) |>
  predict(analysis, new_data = _)
```

Note that any rescaling transformations are automatically handled by `predict()`.

### Single factor

Generate a row for each level of site. Notice that temperature is held at mean value. 
```{r grid-site}
pred_site <- xnew_data(data, site) |>
  predict(analysis, new_data = _)

pred_site

plot_pointrange(pred_site, site) +
  labs(y = "Expected count", x = "Site")
```

### Multiple factors with the hyperprior filling in

When `site` and `annual` are crossed, the full 25-row grid is returned
even though only 22 site×annual cells were observed. Predictions for the
unobserved combinations come from the `z_bSiteAnnual` hyperprior.

```{r grid-site-annual}
xnew_data(data, site, annual) |>
  predict(analysis, new_data = _) |>
  nrow()
```

### `xobs_only()` for observed combinations only

`xobs_only()` restricts the grid to combinations actually present in the
data:

```{r grid-xobs-only}
xnew_data(data, xobs_only(site, annual)) |>
  predict(analysis, new_data = _) |>
  nrow()
```

22 rows instead of 25: exactly the unbalanced combinations from the
design.

### Casting specific factor levels with `xcast()`

To get specific values of factors, use `xcast()`. 
```{r grid-xcast}
xnew_data(
  data,
  xcast(site = "a", treatment = "restored"),
  temperature = 1
) |>
  predict(analysis, new_data = _)
```

`xnew_data(site = "a")` will fail. 

## Scalar quantities from `new_expr`

Pass `new_data = character(0)` to extract a scalar quantity defined in
`new_expr`. `term` selects which scalar:

```{r term-scalar}
predict(analysis, new_data = character(0), term = "eBaseCount")
predict(analysis, new_data = character(0), term = "eRestoredEffect")
```

## Random-effect zeroing

By default (`random_effects = NULL`), a random-effect parameter is zeroed
when its associated factor is held at the first level in `new_data`. This
gives the prediction for the 'typical' (population-level) group.

### Default behaviour

For example, `xnew_data(data, temperature)` leaves both `site` and `annual` at level 1,
so both `z_bSite` and `z_bSiteAnnual` are zeroed, giving the population-level
prediction:

```{r re-default}
xnew_data(data, temperature) |>
  predict(analysis, new_data = _) |>
  head(3)
```

### `random_effects = FALSE`

If `random_effects = FALSE`, zeroing is skipped and predictions are made for the first level of the factor (e.g., for `site` level 1 and `annual` level 1).

```{r re-false}
xnew_data(data, temperature) |>
  predict(analysis, new_data = _, random_effects = FALSE) |>
  head(3)
```

### Named list: zero a subset of random effects

Passing a named list will override the `random_effects` defined in `model()` for more customised behaviour. 

```{r re-named}
xnew_data(data, temperature) |>
  predict(
    analysis,
    new_data = _,
    random_effects = list(z_bSite = "site")
  ) |>
  head(3)
```

## Reference data and effect sizes

`ref_data` must be a flag or a data.frame with one row. When `ref_data` is a one-row data frame, `predict()` computes `ref_fun2(c(ref_draw, new_draw))` per MCMC iteration, where each element is the scalar posterior draw at the reference row and the `new_data` row
respectively. The returned summary describes the posterior of the
*transformed* quantity, not of the raw prediction. The default `ref_fun2`
is `proportional_change2` from the `extras` package (`(new - ref) / ref`), which gives the proportional change of each row in `new_data` with the `ref_data`. When `ref_data` is `TRUE`, a one-row data.frame is automatically generated with all variables held at reference value. 

### Proportional change vs control

```{r ref-prop}
ref_control <- xnew_data(data, xcast(treatment = "control"))

pred_prop <- xnew_data(data, treatment) |>
  predict(analysis, new_data = _, ref_data = ref_control)

pred_prop

plot_pointrange(pred_prop, treatment) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Change vs control", x = "Treatment")
```

The control row sits at exactly 0%. The restored row gives
the proportional change in expected count.

### Custom `ref_fun2`: e.g., absolute difference

Pass any function whose first argument takes a vector of two numbers and
returns a scalar to summarize a different transformation of the two
posteriors. For example, absolute difference reports the effect on the count scale
(restored minus control) rather than as a percentage:

```{r ref-diff}
xnew_data(data, treatment) |>
  predict(
    analysis,
    new_data = _,
    ref_data = ref_control,
    ref_fun2 = function(x) x[2] - x[1]
  )
```

## Predicting for an unobserved level

Zeroing a random effect gives the prediction for the *typical* or *average* group level, e.g., expected count for an average site, with amongst-site variability removed. However, for a *new*, *unobserved* level the uncertainty should be wider to account for amonst-level variation. This is done by drawing a new random effect from its population distribution, which typically will incorporate mean 0 (for random effects centred at 0) and the SD of the random effect, which is estimated. The model's `new_expr` already defines an `unobserved` term to do this:

```r
unobserved[i] <- exp(bIntercept +
  bTreatment[treatment[i]] +
  bTemp * temperature[i] +
  rnorm(1, 0, sSite) +
  rnorm(1, 0, sSiteAnnual))
```

Setting `term = "unobserved"` tells `predict()` to summarize this named
quantity from `new_expr` instead of the default `"prediction"`. Compare the
two scopes side-by-side:

```{r unobserved-vs-typical}
typical <- xnew_data(data, treatment) |>
  predict(analysis, new_data = _) |>
  mutate(scope = "typical site")

unobserved <- xnew_data(data, treatment) |>
  predict(analysis, new_data = _, term = "unobserved") |>
  mutate(scope = "unobserved site")

bind_rows(typical, unobserved)
```

```{r unobserved-plot}
bind_rows(typical, unobserved) |>
  ggplot(aes(treatment, estimate, ymin = lower, ymax = upper, colour = scope)) +
  geom_pointrange(position = position_dodge(width = 0.3)) +
  theme_minimal() +
  labs(y = "Expected count", x = "Treatment", colour = NULL)
```

Compatibility limits are appropriately wider for `unobserved site` as the interval now
reflects amongst-site variation in addition to parameter uncertainty.

## Overriding `new_expr` and injecting constants

`new_expr` can be replaced inline as a string. `new_values` supplies named
scalar constants into that expression's environment. For example, convert
counts to biomass by multiplying through a literature or independently
measured mean fish mass (g), without refitting the count model:

```{r new-expr-override}
xnew_data(data, site) |>
  predict(analysis, new_data = _) |>
  head(2)

predict(
  analysis,
  new_data = xnew_data(data, site),
  new_expr = "
    bSite <- z_bSite * sSite
    for (i in 1:length(site)) {
      prediction[i] <- exp(bIntercept + bSite[site[i]]) * mean_mass_g
    }
  ",
  new_values = list(mean_mass_g = 12)
) |>
  head(2)
```

The second call multiplies the predicted count by an external mean mass of
12 g, giving the posterior of expected biomass per site. Compatibility
limits scale proportionally because the constant has no uncertainty
attached.

## Group-level summaries with `mcmc_derive_data()`

`mcmc_derive_data()` pairs MCMC samples with the per-row data and returns an
`mcmc_data` object. `group_by()` + `summarize()` then aggregates across rows
within each group per MCMC iteration to produce per-group posterior distribution summaries.

### Total expected count per treatment

```{r mcd-total}
mcmc_derive_data(analysis, new_data = data, term = "^eCount$") |>
  group_by(treatment) |>
  summarize() |>
  coef()
```

The default summary function is `sum`, which returns the posterior of *total* abundance per treatment across all rows.

### Per-site mean abundance
Provide a different summarize function, or define a custom one:
```{r mcd-per-site}
mcmc_derive_data(analysis, new_data = data, term = "^eCount$") |>
  group_by(site) |>
  summarize(.fun = mean) |>
  coef()
```

Or a custom function, e.g. coefficient of variation:

```{r mcd-custom-fun}
mcmc_derive_data(analysis, new_data = data, term = "^eCount$") |>
  group_by(treatment) |>
  summarize(.fun = function(x) sd(x) / mean(x)) |>
  coef()
```

## Combining across analyses with `combine_samples()`

`mcmcr::combine_samples()` (with the `mcmc_data` method from `mcmcdata`)
joins two `mcmc_data` objects on shared data columns and applies a function
to align MCMC draws row-by-row. Its natural use is composing quantities
derived from *independent* analyses: for example, multiplying an abundance
posterior by a mass-per-fish posterior to get biomass, where each component
was fitted from a different dataset. Note `mcmc_data` objects are generated from `mcmc_derive_data()`, which is useful for capturing posterior distributions (with associated data) without summarizing, as in `predict()`. 

Assume a separate `mass_analysis` was fitted to size-measurement
data (different observation model, same site factor), and both analyses
expose a per-site quantity:

```{r combine-samples, eval = FALSE}
# Abundance posterior per site, from the counts model fitted above
abundance <- mcmc_derive_data(
  analysis,
  new_data = xnew_data(data, site),
  term = "^eCount$"
)

# Mass-per-fish posterior per site, from a separate analysis (sketch)
mass_per_fish <- mcmc_derive_data(
  mass_analysis,
  new_data = xnew_data(mass_data, site),
  term = "^eMass$"
)

# Per-site biomass posterior: abundance * mass-per-fish
biomass <- mcmcr::combine_samples(
  abundance,
  mass_per_fish,
  by = "site",
  fun = prod
)

coef(biomass)
```

`by` names the data columns to join on (here `site`); rows present in both
inputs are paired. `fun` is applied to the matched MCMC draws, so `prod`
gives the row-wise product across iterations. The result is an `mcmc_data`
with one row per joined key and a propagated posterior in `biomass`.

For the two analyses to combine:

* **MCMC dimensions must match.** Both fits need the same number of chains
  and iterations (e.g. both 3 chains × 500 saved draws). `combine_samples()`
  pairs draws position-by-position; mismatched dimensions error out.
* **`by` columns must exist in both `$data` slots.** Rows are paired by an
  inner join, so non-matching rows are dropped silently. Make sure factor
  levels and column types line up.

Within a single analysis, prefer expressing the composition directly in
`new_expr`; `combine_samples()` is most useful when the components come from
independently fitted models whose draws need joining.

## Scalar derived quantities with `mcmc_derive()`

`mcmc_derive()` returns the raw MCMC samples for one or more scalar terms in
`new_expr`. As with `mcmc_derive_data()`, this is useful when you want to compute posteriors without summarizing to a point estimate and compatibility limits. 

### Pull related scalars by regex

Unlike `predict()`, the `term` argument in `mcmc_derive()` can accept a regular expression. For example, both `eBaseCount` and `eRestoredEffect` are extracted in one call:

```{r mcmc-derive-regex}
scalars <- mcmc_derive(
  analysis,
  new_data = character(0),
  term = "^(eBaseCount|eRestoredEffect)$"
)

coef(scalars)
```

`coef()` is run on the `mcmc_derive()` output to get point estimte with compatibility limits. 

### Arithmetic on `mcmcr` objects propagates uncertainty

Multiplying `mcmcr` objects acts element-wise across MCMC iterations:

```{r mcmc-derive-arithmetic}
eExpectedRestored <- scalars$eBaseCount * scalars$eRestoredEffect
class(eExpectedRestored)
coef(eExpectedRestored)
```

## Cheat sheet

| Goal | Tool |
|---|---|
| Posterior summary at new covariate values | `predict()` |
| Posterior summary for a scalar derived quantity | `predict()` with `new_data = character(0)` and `term` specified |
| Prediction for an unobserved factor level (includes amongst-level variation) | `new_expr` term with `rnorm(1, 0, s*)` for the random effect |
| Effect size (proportional, absolute) | `predict()` with `ref_data` + `ref_fun2` |
| Group-level aggregates (sum, mean, custom) per MCMC iteration | `mcmc_derive_data()` + `group_by()` + `summarize()` |
| Combine quantities derived from independent analyses | `mcmcr::combine_samples()` |
| Scalar parameter posteriors and arithmetic | `mcmc_derive()` |

See `?predict.mb_analysis`, `?mcmc_derive.mb_analysis`,
`?mcmc_derive_data.mb_analysis`, and `?mcmcr::combine_samples` for full
argument documentation.


