---
title: "Prediction with predict() and Related Functions"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  cache.lazy = FALSE,
  fig.width = 7,
  fig.height = 4
)
set.seed(42)
```

`predict.mb_analysis()` summarizes the posterior predictive distribution of a
fitted Bayesian model at new covariate values, returning a tidy data frame
with point estimates and compatibility limits. This vignette covers:

* [Covariate grids](#covariate-grids-with-the-newdata-helpers) built with the `newdata` helpers.
* [Scalar quantities from `new_expr`](#scalar-quantities-from-new_expr).
* [Random-effect zeroing](#random-effect-zeroing) control.
* [Reference data and effect sizes](#reference-data-and-effect-sizes), e.g., for proportional change.
* [Predicting for an unobserved level](#predicting-for-an-unobserved-level) to include amongst-level variation.
* [Overriding `new_expr` and injecting constants](#overriding-new_expr-and-injecting-constants) for ad-hoc derived quantities.
* [Group-level summaries with `mcmc_derive_data()`](#group-level-summaries-with-mcmc_derive_data).
* [Combining across analyses with `combine_samples()`](#combining-across-analyses-with-combine_samples) for joining and summarizing samples from independent fits.
* [Scalar derived quantities with `mcmc_derive()`](#scalar-derived-quantities-with-mcmc_derive) and arithmetic on raw `mcmcr` posteriors.
* [Cheat sheet](#cheat-sheet) mapping common goals to tools.

```{r libraries, message = FALSE}
library(embr)
library(smbr2)
library(newdata)
library(mcmcr)
library(mcmcderive)
library(dplyr)
library(ggplot2)
```

```{r plot-helpers, include = FALSE}
plot_pointrange <- function(df, x) {
  ggplot(df, aes({{ x }}, estimate, ymin = lower, ymax = upper)) +
    geom_pointrange() +
    theme_minimal()
}

plot_ribbon <- function(df, x) {
  ggplot(df, aes({{ x }}, estimate, ymin = lower, ymax = upper)) +
    geom_ribbon(alpha = 0.2) +
    geom_line() +
    theme_minimal()
}
```

## Setup: fit demo model
A simple example fish count model is fit with Stan (via cmdstanr and smbr2). The simulated data has 5 sites × 5 years × 2 treatments, with an unbalanced design (i.e., not all sites have all years with data). 

```{r simulate, include = FALSE}
sites <- letters[1:5]
years <- 1:5
treatments <- c("control", "restored")

design <- expand.grid(
  site = sites,
  annual = years,
  treatment = treatments,
  rep = 1:3,
  stringsAsFactors = FALSE
) |>
  filter(!(site == "e" & annual %in% c(1, 2, 3))) |>
  mutate(
    site = factor(site, levels = sites),
    annual = factor(annual, levels = years),
    treatment = factor(treatment, levels = treatments),
    temperature = rnorm(n(), mean = 12, sd = 2)
  )

re_site <- setNames(rnorm(length(sites), 0, 0.4), sites)
re_annual <- setNames(rnorm(length(years), 0, 0.2), as.character(years))

data <- design |>
  mutate(
    temp_c = (temperature - mean(temperature)) / sd(temperature),
    log_mu = log(40) +
      0.5 * (treatment == "restored") +
      0.15 * temp_c +
      re_site[as.character(site)] +
      re_annual[as.character(annual)],
    count = as.integer(rnbinom(n(), mu = exp(log_mu), size = 5))
  ) |>
  select(count, site, annual, treatment, temperature)
```

```{r stan-code}
count_code <- "
data {
  int<lower=2> nObs;
  int<lower=1> nsite;
  int<lower=1> nannual;
  int<lower=1> ntreatment;
  array[nObs] int<lower=1> site;
  array[nObs] int<lower=1> annual;
  array[nObs] int<lower=1> treatment;
  vector[nObs] temperature;
  array[nObs] int<lower=0> count;
}
parameters {
  real bIntercept;
  vector[ntreatment - 1] bTreatment_dev;
  real bTemp;
  real<lower=0> sSite;
  real<lower=0> sAnnual;
  real<lower=0> bPhi;
  vector[nsite] z_bSite;
  vector[nannual] z_bAnnual;
}
model {
  vector[nsite] bSite = z_bSite * sSite;
  vector[nannual] bAnnual = z_bAnnual * sAnnual;
  vector[ntreatment] bTreatment;
  bTreatment[1] = 0;
  for (k in 2:ntreatment) bTreatment[k] = bTreatment_dev[k - 1];

  vector[nObs] log_eCount;
  for (i in 1:nObs) {
    log_eCount[i] = bIntercept + bTreatment[treatment[i]] +
                    bTemp * temperature[i] + bSite[site[i]] +
                    bAnnual[annual[i]];
  }

  bIntercept ~ normal(0, 5);
  bTreatment_dev ~ normal(0, 5);
  bTemp ~ normal(0, 5);
  sSite ~ exponential(1);
  sAnnual ~ exponential(1);
  bPhi ~ gamma(2, 0.5);
  to_vector(z_bSite) ~ std_normal();
  to_vector(z_bAnnual) ~ std_normal();

  count ~ neg_binomial_2_log(log_eCount, bPhi);
}
"

count_model <- model(
  code = count_code,
  new_expr = {
    bSite <- z_bSite * sSite
    bAnnual <- z_bAnnual * sAnnual
    bTreatment <- c(0, bTreatment_dev)
    eBaseCount <- exp(bIntercept)
    eRestoredEffect <- exp(bTreatment_dev[1])
    for (i in 1:nObs) {
      log(eCount[i]) <- bIntercept +
        bTreatment[treatment[i]] +
        bTemp * temperature[i] +
        bSite[site[i]] +
        bAnnual[annual[i]]
      prediction[i] <- eCount[i]
      fit[i] <- eCount[i]
      unobserved[i] <- exp(
        bIntercept +
          bTreatment[treatment[i]] +
          bTemp * temperature[i] +
          rnorm(1, 0, sSite) +
          rnorm(1, 0, sAnnual)
      )
      log_lik[i] <- log_lik_neg_binom(count[i], eCount[i], bPhi)
    }
  },
  new_expr_vec = TRUE,
  select_data = list(
    count = 1L,
    site = factor("a"),
    annual = factor("a"),
    treatment = factor("a"),
    `temperature*` = 1
  ),
  random_effects = list(
    z_bSite = "site",
    z_bAnnual = "annual"
  )
)
```

```{r fit, message = FALSE}
analysis <- analyse(
  count_model,
  data = data,
  stan_engine = "cmdstan-mcmc",
  nchains = 3L,
  niters = 500L,
  niters_warmup = 500L,
  seed = 42L,
  parallel = TRUE,
  quiet = TRUE,
  beep = FALSE
)

coef(analysis, include_constant = FALSE, simplify = TRUE, directional_information = FALSE) |>
  mutate(across(estimate:svalue, ~ signif(.x, 3)))
```

## Default prediction

With no `new_data`, `predict()` returns predicted values for every row of the
analysis dataset. This is pulled from the `prediction[i]` line in `new_expr`:

```{r default-predict}
predict(analysis) |> head()
```

## Covariate grids with the `newdata` helpers

Covariate grids can be generated with `newdata::xnew_data()`. Covariates not specified are held at a reference value: mean for continuous, first level for factors. For categorical fixed effects this is the literal first level; for random-effect factors at the first level, that random effect is zeroed by default — see *Random-effect zeroing* below.

### Continuous covariate
By default, specifying the name of a continuous covariate will generate a sequence from min to max, with length 30. Notice factors (e.g., site) are held at the first level. 

```{r grid-temperature}
temp_data <- xnew_data(data, temperature)
head(temp_data)
nrow(temp_data)

pred_temp <- predict(analysis, new_data = temp_data)

head(pred_temp)

plot_ribbon(pred_temp, temperature) +
  labs(y = "Expected count", x = "Temperature (scaled)")
```

### Specific values via `xnew_seq()` or with `=`

```{r grid-seq}
xnew_data(data, xnew_seq(temperature, length_out = 5)) |>
  predict(analysis, new_data = _)
```

```{r grid-eq}
xnew_data(data, temperature = 10) |>
  predict(analysis, new_data = _)
```

Note that any rescaling transformations are automatically handled by `predict()`.

### Single factor

Generate a row for each level of site. Notice that temperature is held at mean value. 
```{r grid-site}
pred_site <- xnew_data(data, site) |>
  predict(analysis, new_data = _)

pred_site

plot_pointrange(pred_site, site) +
  labs(y = "Expected count", x = "Site")
```

### `xobs_only()` for observed combinations only

By default `xnew_data(data, site, annual)` returns the full 25-row crossed
grid:

```{r grid-site-annual}
xnew_data(data, site, annual) |>
  predict(analysis, new_data = _) |>
  nrow()
```

If you only want predictions for the (`site`, `annual`) combinations
actually observed in the data, wrap them in `xobs_only()`:

```{r grid-xobs-only}
xnew_data(data, xobs_only(site, annual)) |>
  predict(analysis, new_data = _) |>
  nrow()
```

22 rows instead of 25: the unbalanced combinations from the design (site
`e` is missing from years 1, 2, and 3).

### Casting specific factor levels with `xcast()`

To get specific values of factors, use `xcast()`. 
```{r grid-xcast}
xnew_data(
  data,
  xcast(site = "a", treatment = "restored"),
  temperature = 1
) |>
  predict(analysis, new_data = _)
```

Note that `xnew_data(data, site = "a")` will not work: bare `=` assignment is for *continuous* covariates (e.g., `temperature = 1`). To fix a factor at a specific level, wrap it in `xcast()`.

## Scalar quantities from `new_expr`

Pass `new_data = character(0)` to extract a scalar quantity defined in
`new_expr`. `term` selects which scalar:

```{r term-scalar}
predict(analysis, new_data = character(0), term = "eBaseCount")
predict(analysis, new_data = character(0), term = "eRestoredEffect")
```

## Random-effect zeroing

By default (`random_effects = NULL`), a random-effect parameter is zeroed
when its associated factor is held at the first level in `new_data`. This
gives the prediction for the 'typical' (population-level) group.

### Default behaviour

For example, `xnew_data(data, temperature)` leaves both `site` and `annual`
at level 1, so both `z_bSite` and `z_bAnnual` are zeroed, giving the
population-level prediction:

```{r re-default}
xnew_data(data, temperature) |>
  predict(analysis, new_data = _) |>
  head(3)
```

### `random_effects = FALSE`

If `random_effects = FALSE`, zeroing is skipped and predictions use the literal first level of each random-effect factor (here, `site = "a"` and `annual = 1`).

```{r re-false}
xnew_data(data, temperature) |>
  predict(analysis, new_data = _, random_effects = FALSE) |>
  head(3)
```

### Named list: zero a subset of random effects

Passing a named list overrides the `random_effects` defined in `model()`,
restricting the set of parameters eligible for zeroing to those listed.
Parameters not in the list are kept at their literal value for the first
level of their associated factor in `new_data`. This is useful when you
want to marginalize over one random effect while keeping another at a
specific level.

```{r re-named}
xnew_data(data, temperature) |>
  predict(
    analysis,
    new_data = _,
    random_effects = list(z_bSite = "site")
  ) |>
  head(3)
```

For the call above (with `site` and `annual` both at level 1 in
`new_data`):

- `z_bSite` is listed and `site` is at level 1, so `z_bSite` is zeroed
  (population-mean site, i.e. averaged across sites).
- `z_bAnnual` is *not* listed, so it is kept at its literal value for
  `annual = 1` — the specific year-1 effect.

This gives the prediction averaged across sites but anchored to year 1
specifically.


## Reference data and effect sizes

`ref_data` must be a flag or a data.frame with one row. When `ref_data` is a one-row data frame, `predict()` computes `ref_fun2(c(ref_draw, new_draw))` per MCMC iteration, where each element is the scalar posterior draw at the reference row and the `new_data` row
respectively. The returned summary describes the posterior of the
*transformed* quantity, not of the raw prediction. The default `ref_fun2`
is `proportional_change2` from the `extras` package (`(new - ref) / ref`), which gives the proportional change of each row in `new_data` with the `ref_data`. When `ref_data` is `TRUE`, a one-row data.frame is automatically generated with all variables held at reference value. 

### Proportional change vs control

```{r ref-prop}
ref_control <- xnew_data(data, xcast(treatment = "control"))

pred_prop <- xnew_data(data, treatment) |>
  predict(analysis, new_data = _, ref_data = ref_control)

pred_prop

plot_pointrange(pred_prop, treatment) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Change vs control", x = "Treatment")
```

The control row sits at exactly 0%. The restored row gives
the proportional change in expected count.

### Custom `ref_fun2`: e.g., absolute difference

Pass any function whose first argument takes a vector of two numbers and
returns a scalar to summarize a different transformation of the two
posteriors. For example, absolute difference reports the effect on the count scale
(restored minus control) rather than as a percentage:

```{r ref-diff}
pred_diff <- xnew_data(data, treatment) |>
  predict(
    analysis,
    new_data = _,
    ref_data = ref_control,
    ref_fun2 = function(x) x[2] - x[1]
  )

pred_diff

plot_pointrange(pred_diff, treatment) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(y = "Change vs control (count)", x = "Treatment")
```

## Predicting for an unobserved level

Zeroing a random effect gives the prediction for the *typical* or *average* group level, e.g., expected count for an average site, with amongst-site variability removed. However, for a *new*, *unobserved* level the uncertainty should be wider to account for amongst-level variation. This is done by drawing a new random effect from its population distribution, which typically will incorporate mean 0 (for random effects centred at 0) and the SD of the random effect, which is estimated. The model's `new_expr` already defines an `unobserved` term to do this:

```r
unobserved[i] <- exp(bIntercept +
  bTreatment[treatment[i]] +
  bTemp * temperature[i] +
  rnorm(1, 0, sSite) +
  rnorm(1, 0, sAnnual))
```

Setting `term = "unobserved"` tells `predict()` to summarize this named
quantity from `new_expr` instead of the default `"prediction"`. Compare the
two scopes side-by-side:

```{r unobserved-vs-typical}
typical <- xnew_data(data, treatment) |>
  predict(analysis, new_data = _) |>
  mutate(scope = "typical site")

unobserved <- xnew_data(data, treatment) |>
  predict(analysis, new_data = _, term = "unobserved") |>
  mutate(scope = "unobserved site")

bind_rows(typical, unobserved)
```

```{r unobserved-plot}
bind_rows(typical, unobserved) |>
  ggplot(aes(treatment, estimate, ymin = lower, ymax = upper, colour = scope)) +
  geom_pointrange(position = position_dodge(width = 0.3)) +
  theme_minimal() +
  labs(y = "Expected count", x = "Treatment", colour = NULL)
```

Compatibility limits are wider for `unobserved site`, as the interval now
reflects amongst-site variation in addition to parameter uncertainty.

## Overriding `new_expr` and injecting constants

`new_expr` can be replaced inline as a string. `new_values` supplies named
scalar constants into that expression's environment. For example, convert
counts to biomass by multiplying through a literature or independently
measured mean fish mass (g), without refitting the count model:

```{r new-expr-override}
xnew_data(data, site) |>
  predict(analysis, new_data = _) |>
  head(2)

predict(
  analysis,
  new_data = xnew_data(data, site),
  new_expr = "
    bSite <- z_bSite * sSite
    for (i in 1:length(site)) {
      prediction[i] <- exp(bIntercept + bSite[site[i]]) * mean_mass_g
    }
  ",
  new_values = list(mean_mass_g = 12)
) |>
  head(2)
```

The second call multiplies the predicted count by an external mean mass of
12 g, giving the posterior of expected biomass per site. Compatibility
limits scale proportionally because the constant has no uncertainty
attached.

## Group-level summaries with `mcmc_derive_data()`

`mcmc_derive_data()` pairs MCMC samples with the per-row data and returns an
`mcmc_data` object. `group_by()` + `summarize()` then aggregates across rows
within each group per MCMC iteration to produce per-group posterior distribution summaries.

### Total expected count per treatment

```{r mcd-total}
mcmc_derive_data(analysis, new_data = data, term = "^eCount$") |>
  group_by(treatment) |>
  summarize() |>
  coef()
```

The default summary function is `sum`, which returns the posterior of *total* abundance per treatment across all rows.

### Custom summary function

Provide any function via `.fun`. For example, the coefficient of variation
of expected count across rows within each treatment:

```{r mcd-custom-fun}
mcmc_derive_data(analysis, new_data = data, term = "^eCount$") |>
  group_by(treatment) |>
  summarize(.fun = function(x) sd(x) / mean(x)) |>
  coef()
```


## Combining across analyses with `combine_samples()`

`mcmc_data` objects come from `mcmc_derive_data()`; they hold the raw
per-row MCMC samples together with the associated data, without the
point-estimate-and-limits summarisation that `predict()` performs.

`mcmcr::combine_samples()` (with the `mcmc_data` method from `mcmcdata`)
joins two `mcmc_data` objects on shared data columns and applies a function
to align MCMC draws row-by-row. Its natural use is composing quantities
derived from *independent* analyses: for example, multiplying an abundance
posterior by a mass-per-fish posterior to get biomass, where each component
was fitted from a different dataset.

Assume a separate `mass_analysis` was fitted to size-measurement
data (different observation model, same site factor), and both analyses
expose a per-site quantity:

```{r combine-samples, eval = FALSE}
# Abundance posterior per site, from the counts model fitted above
abundance <- mcmc_derive_data(
  analysis,
  new_data = xnew_data(data, site),
  term = "^eCount$"
)

# Mass-per-fish posterior per site, from a separate analysis (sketch)
mass_per_fish <- mcmc_derive_data(
  mass_analysis,
  new_data = xnew_data(mass_data, site),
  term = "^eMass$"
)

# Per-site biomass posterior: abundance * mass-per-fish
biomass <- mcmcr::combine_samples(
  abundance,
  mass_per_fish,
  by = "site",
  fun = prod
)
```

`by` names the data columns to join on (here `site`); rows present in both
inputs are paired. `fun` is applied to the matched MCMC draws, so `prod`
gives the row-wise product across iterations. The result is an `mcmc_data`
with one row per joined key and a propagated posterior in `biomass`.

For the two analyses to combine:

* **MCMC dimensions must match.** Both fits need the same number of chains
  and iterations (e.g. both 3 chains × 500 saved draws). `combine_samples()`
  pairs draws position-by-position; mismatched dimensions error out.
* **`by` columns must exist in both `$data` slots.** Rows are paired by an
  inner join, so non-matching rows are dropped silently. Make sure factor
  levels and column types line up.

Within a single analysis, prefer expressing the composition directly in
`new_expr`; `combine_samples()` is most useful when the components come from
independently fitted models whose draws need joining.

## Scalar derived quantities with `mcmc_derive()`

`mcmc_derive()` returns the raw MCMC samples for one or more scalar terms
in `new_expr`. As with `mcmc_derive_data()`, this is useful when you want
to work with the raw MCMC samples directly rather than `predict()`'s
estimate-and-limits summary.

### Pull related scalars by regex

Unlike `predict()`, the `term` argument in `mcmc_derive()` can accept a regular expression. For example, both `eBaseCount` and `eRestoredEffect` are extracted in one call:

```{r mcmc-derive-regex}
scalars <- mcmc_derive(
  analysis,
  new_data = character(0),
  term = "^(eBaseCount|eRestoredEffect)$"
)

coef(scalars)
```

`coef()` is run on the `mcmc_derive()` output to get point estimates with compatibility limits.

### Arithmetic on `mcmcr` objects

Quantities you anticipate are usually cleanest to define directly in
`new_expr` and pull with `predict()`. Raw `mcmcr` arithmetic earns its
keep when the question requires *composing* outputs of different
post-fit operations — for example, combining a group-aggregated quantity
from `mcmc_derive_data()` with a scalar from `mcmc_derive()`, or
producing summaries `predict()` cannot (probability statements, custom
posterior quantiles).

First, get totals per treatment, then extract each row as raw MCMC draws:

```{r mcmc-derive-totals}
totals <- mcmc_derive_data(analysis, new_data = data, term = "^eCount$") |>
  group_by(treatment) |>
  summarize()

restored_total <- as.mcmcr(filter(totals, treatment == "restored"))[[1]]
control_total <- as.mcmcr(filter(totals, treatment == "control"))[[1]]
```

The posterior of the absolute difference in total expected catch
(restored minus control), summed over the observed design — a count-scale
effect size that `eRestoredEffect` (a multiplicative ratio) doesn't
quantify:

```{r mcmc-derive-diff}
extra_catch <- restored_total - control_total
coef(extra_catch)
```

And the posterior probability that restoration adds at least 2000 fish
in total — a question `predict()` cannot answer:

```{r mcmc-derive-prob}
mean(extra_catch >= 2000)
```

## Cheat sheet

| Goal | Tool |
|---|---|
| Posterior summary at new covariate values | `predict()` |
| Posterior summary for a scalar derived quantity | `predict()` with `new_data = character(0)` and `term` specified |
| Prediction for an unobserved factor level (includes amongst-level variation) | `new_expr` term with `rnorm(1, 0, s*)` for the random effect |
| Effect size (proportional, absolute) | `predict()` with `ref_data` + `ref_fun2` |
| Group-level aggregates (sum, mean, custom) per MCMC iteration | `mcmc_derive_data()` + `group_by()` + `summarize()` |
| Combine quantities derived from independent analyses | `mcmcr::combine_samples()` |
| Scalar parameter posteriors and arithmetic | `mcmc_derive()` |

See `?predict.mb_analysis`, `?mcmc_derive.mb_analysis`,
`?mcmc_derive_data.mb_analysis`, and `?mcmcr::combine_samples` for full
argument documentation.


