---
title: "Prediction with predict() and Related Functions"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  cache.lazy = FALSE,
  fig.width = 7,
  fig.height = 4
)
set.seed(42)
```

`predict.mb_analysis()` summarises the posterior predictive distribution of a
fitted Bayesian model at new covariate values, returning a tidy data frame
with point estimates and compatibility limits. This vignette covers:

* [Covariate grids](#covariate-grids-with-the-newdata-helpers) built with the `newdata` helpers.
* [Scalar quantities from `new_expr`](#scalar-quantities-from-new_expr).
* [Reference data and effect sizes](#reference-data-and-effect-sizes) for proportional and absolute effects.
* [Predicting for an unobserved site](#predicting-for-an-unobserved-site) to include amongst-level variation.
* [Overriding `new_expr` and injecting constants](#overriding-new_expr-and-injecting-constants) for ad-hoc derived quantities.
* [Group-level summaries with `mcmc_derive_data()`](#group-level-summaries-with-mcmc_derive_data) for per-group aggregates.
* [Combining across analyses with `combine_samples()`](#combining-across-analyses-with-combine_samples) for joining samples from independent fits.
* [Scalar derived quantities with `mcmc_derive()`](#scalar-derived-quantities-with-mcmc_derive) and arithmetic on `mcmcr` posteriors.
* [Random-effect zeroing](#random-effect-zeroing) for controlling which `z_b*` parameters collapse to the population level.
* [Cheat sheet](#cheat-sheet) mapping common goals to tools.

```{r libraries, message = FALSE}
library(embr)
library(smbr2)
library(newdata)
library(mcmcr)
library(mcmcderive)
library(dplyr)
library(ggplot2)
```

```{r plot-helpers, include = FALSE}
plot_pointrange <- function(df, x) {
  ggplot(df, aes({{ x }}, estimate, ymin = lower, ymax = upper)) +
    geom_pointrange() +
    theme_minimal()
}

plot_ribbon <- function(df, x) {
  ggplot(df, aes({{ x }}, estimate, ymin = lower, ymax = upper)) +
    geom_ribbon(alpha = 0.2) +
    geom_line() +
    theme_minimal()
}
```

## Setup: fit demo model
A simple demo fish count model is fit with Stan (via cmdstanr and smbr2). The simulated data has 5 sites × 5 years × 2 treatments, with an unbalanced design (i.e., not all sites have all years with data). 

```{r simulate, include = FALSE}
sites <- letters[1:5]
years <- 1:5
treatments <- c("control", "restored")

design <- expand.grid(
  site = sites,
  annual = years,
  treatment = treatments,
  rep = 1:3,
  stringsAsFactors = FALSE
) |>
  filter(!(site == "e" & annual %in% c(1, 2, 3))) |>
  mutate(
    site = factor(site, levels = sites),
    annual = factor(annual, levels = years),
    treatment = factor(treatment, levels = treatments),
    temperature = rnorm(n(), mean = 12, sd = 2)
  )

re_site <- setNames(rnorm(length(sites), 0, 0.4), sites)
re_site_annual <- matrix(
  rnorm(length(sites) * length(years), 0, 0.2),
  nrow = length(sites),
  dimnames = list(sites, years)
)

data <- design |>
  mutate(
    temp_c = (temperature - mean(temperature)) / sd(temperature),
    log_mu = log(40) +
      0.5 * (treatment == "restored") +
      0.15 * temp_c +
      re_site[as.character(site)] +
      mapply(
        function(s, a) re_site_annual[s, a],
        as.character(site),
        as.character(annual)
      ),
    count = as.integer(rnbinom(n(), mu = exp(log_mu), size = 5))
  ) |>
  select(count, site, annual, treatment, temperature)
```

```{r stan-code}
count_code <- "
data {
  int<lower=2> nObs;
  int<lower=1> nsite;
  int<lower=1> nannual;
  int<lower=1> ntreatment;
  array[nObs] int<lower=1> site;
  array[nObs] int<lower=1> annual;
  array[nObs] int<lower=1> treatment;
  vector[nObs] temperature;
  array[nObs] int<lower=0> count;
}
parameters {
  real bIntercept;
  vector[ntreatment - 1] bTreatment_dev;
  real bTemp;
  real<lower=0> sSite;
  real<lower=0> sSiteAnnual;
  real<lower=0> bPhi;
  vector[nsite] z_bSite;
  matrix[nsite, nannual] z_bSiteAnnual;
}
model {
  vector[nsite] bSite = z_bSite * sSite;
  matrix[nsite, nannual] bSiteAnnual = z_bSiteAnnual * sSiteAnnual;
  vector[ntreatment] bTreatment;
  bTreatment[1] = 0;
  for (k in 2:ntreatment) bTreatment[k] = bTreatment_dev[k - 1];

  vector[nObs] log_eCount;
  for (i in 1:nObs) {
    log_eCount[i] = bIntercept + bTreatment[treatment[i]] +
                    bTemp * temperature[i] + bSite[site[i]] +
                    bSiteAnnual[site[i], annual[i]];
  }

  bIntercept ~ normal(0, 5);
  bTreatment_dev ~ normal(0, 5);
  bTemp ~ normal(0, 5);
  sSite ~ exponential(1);
  sSiteAnnual ~ exponential(1);
  bPhi ~ gamma(2, 0.5);
  to_vector(z_bSite) ~ std_normal();
  to_vector(z_bSiteAnnual) ~ std_normal();

  count ~ neg_binomial_2_log(log_eCount, bPhi);
}
"

count_model <- model(
  code = count_code,
  new_expr = {
    bSite <- z_bSite * sSite
    bSiteAnnual <- z_bSiteAnnual * sSiteAnnual
    bTreatment <- c(0, bTreatment_dev)
    eBaseCount <- exp(bIntercept)
    eRestoredEffect <- exp(bTreatment_dev[1])
    for (i in 1:nObs) {
      log(eCount[i]) <- bIntercept +
        bTreatment[treatment[i]] +
        bTemp * temperature[i] +
        bSite[site[i]] +
        bSiteAnnual[site[i], annual[i]]
      prediction[i] <- eCount[i]
      fit[i] <- eCount[i]
      unobserved[i] <- exp(
        bIntercept +
          bTreatment[treatment[i]] +
          bTemp * temperature[i] +
          rnorm(1, 0, sSite) +
          rnorm(1, 0, sSiteAnnual)
      )
      log_lik[i] <- log_lik_neg_binom(count[i], eCount[i], bPhi)
    }
  },
  new_expr_vec = TRUE,
  select_data = list(
    count = 1L,
    site = factor("a"),
    annual = factor("a"),
    treatment = factor("a"),
    `temperature*` = 1
  ),
  random_effects = list(
    z_bSite = "site",
    z_bSiteAnnual = c("site", "annual")
  )
)
```

```{r fit, message = FALSE}
analysis <- analyse(
  count_model,
  data = data,
  stan_engine = "cmdstan-mcmc",
  nchains = 3L,
  niters = 500L,
  niters_warmup = 500L,
  seed = 42L,
  parallel = TRUE,
  quiet = TRUE,
  beep = FALSE
)

coef(analysis, include_constant = FALSE, simplify = TRUE) |>
  mutate(across(estimate:svalue, ~ signif(.x, 3)))
```

## Default prediction

With no `new_data`, `predict()` returns predicted values for every row of the
analysis dataset:

```{r default-predict}
predict(analysis) |> head()
```

## Covariate grids with the `newdata` helpers

Covariates not specified in `xnew_data()` are held at a reference value:
mean for continuous, first level for factors.

### Continuous covariate
By default, specifying the name of a continuous covariate will generate a sequence from min to max with length 30.

```{r grid-temperature}
temp_data <- xnew_data(data, temperature)
head(temp_data)
nrow(temp_data)

pred_temp <- predict(analysis, new_data = temp_data)

head(pred_temp)

plot_ribbon(pred_temp, temperature) +
  labs(y = "Expected count", x = "Temperature (scaled)")
```

### Specific values via `xnew_seq()` or with `=`

```{r grid-seq}
xnew_data(data, xnew_seq(temperature, length_out = 5)) |>
  predict(analysis, new_data = _)
```

```{r grid-eq}
xnew_data(data, temperature = 10) |>
  predict(analysis, new_data = _)
```

Note that any required transformations due to centering or scaling are automatically handled by `predict()`.

### Single factor

```{r grid-site}
pred_site <- xnew_data(data, site) |>
  predict(analysis, new_data = _)

pred_site

plot_pointrange(pred_site, site) +
  labs(y = "Expected count", x = "Site")
```

### Multiple factors with the hyperprior filling in

When `site` and `annual` are crossed, the full 25-row grid is returned
even though only 22 site×annual cells were observed. Predictions for the
unobserved combinations come from the `z_bSiteAnnual` hyperprior.

```{r grid-site-annual}
xnew_data(data, site, annual) |>
  predict(analysis, new_data = _) |>
  nrow()
```

### `xobs_only()` for observed combinations only

`xobs_only()` restricts the grid to combinations actually present in the
data:

```{r grid-xobs-only}
xnew_data(data, xobs_only(site, annual)) |>
  predict(analysis, new_data = _) |>
  nrow()
```

22 rows instead of 25: exactly the unbalanced combinations from the
design.

### Casting specific factor levels with `xcast()`

```{r grid-xcast}
xnew_data(
  data,
  xcast(site = "a", treatment = "restored"),
  temperature = 1
) |>
  predict(analysis, new_data = _)
```

## Scalar quantities from `new_expr`

Pass `new_data = character(0)` to extract a scalar quantity defined in
`new_expr`. `term` selects which scalar:

```{r term-scalar}
predict(analysis, new_data = character(0), term = "eBaseCount")
predict(analysis, new_data = character(0), term = "eRestoredEffect")
```

`eBaseCount` is the expected count at reference temperature and the control
treatment (population-level). `eRestoredEffect` is the multiplicative effect
of the restoration treatment.

## Reference data and effect sizes

When `ref_data` is a 1-row data frame, `predict()` computes
`ref_fun2(c(ref_iter, new_iter))` per MCMC iteration and summarises the
result. The returned summary describes the posterior of the *transformed*
quantity, not of the raw prediction. The default `ref_fun2` is
`proportional_change2` from the `extras` package (`(new - ref) / ref`).

### Proportional change vs control

```{r ref-prop}
ref_control <- xnew_data(data, xcast(treatment = "control"))

pred_prop <- xnew_data(data, treatment) |>
  predict(analysis, new_data = _, ref_data = ref_control)

pred_prop

plot_pointrange(pred_prop, treatment) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Change vs control", x = "Treatment")
```

The control row sits at exactly 0% by construction. The restored row gives
the proportional change in expected count.

### Custom `ref_fun2`: absolute difference

Pass any two-argument function `f(x)` (where `x` is the vector
`c(ref_iter, new_iter)`) to summarise a different transformation of the
two posteriors. Absolute difference reports the effect on the count scale
(restored minus control) rather than as a percentage, which is often easier
to interpret directly:

```{r ref-diff}
xnew_data(data, treatment) |>
  predict(
    analysis,
    new_data = _,
    ref_data = ref_control,
    ref_fun2 = function(x) x[2] - x[1]
  )
```

## Predicting for an unobserved site

Zeroing a random effect (the default for the omitted-factor case below)
gives the prediction for the *typical* or *average* group, e.g., expected count for an average site, with amongst-site variability collapsed away. To predict
for a *new* site that wasn't in the data with appropriate uncertainty, draw a fresh random effect from its population distribution. The model's `new_expr` already defines an `unobserved` term to do this:

```r
unobserved[i] <- exp(bIntercept +
  bTreatment[treatment[i]] +
  bTemp * temperature[i] +
  rnorm(1, 0, sSite) +
  rnorm(1, 0, sSiteAnnual))
```

Each MCMC iteration draws fresh site and site×annual deviations from the
posterior of `sSite` and `sSiteAnnual`, propagating amongst-site
uncertainty into the prediction.

Setting `term = "unobserved"` tells `predict()` to summarise this named
quantity from `new_expr` instead of the default `"prediction"`. Compare the
two scopes side-by-side:

```{r unobserved-vs-typical}
typical <- xnew_data(data, treatment) |>
  predict(analysis, new_data = _) |>
  mutate(scope = "typical site")

unobserved <- xnew_data(data, treatment) |>
  predict(analysis, new_data = _, term = "unobserved") |>
  mutate(scope = "unobserved site")

bind_rows(typical, unobserved)
```

```{r unobserved-plot}
bind_rows(typical, unobserved) |>
  ggplot(aes(treatment, estimate, ymin = lower, ymax = upper, colour = scope)) +
  geom_pointrange(position = position_dodge(width = 0.3)) +
  theme_minimal() +
  labs(y = "Expected count", x = "Treatment", colour = NULL)
```

Compatibility limits are wider for `unobserved site`; the interval now
reflects amongst-site variation in addition to parameter uncertainty.

## Overriding `new_expr` and injecting constants

`new_expr` can be replaced inline as a string. `new_values` supplies named
scalar constants into that expression's environment. For example, convert
counts to biomass by multiplying through a literature or independently
measured mean fish mass (g), without refitting the count model:

```{r new-expr-override}
xnew_data(data, site) |>
  predict(analysis, new_data = _) |>
  head(2)

predict(
  analysis,
  new_data = xnew_data(data, site),
  new_expr = "
    bSite <- z_bSite * sSite
    for (i in 1:length(site)) {
      prediction[i] <- exp(bIntercept + bSite[site[i]]) * mean_mass_g
    }
  ",
  new_values = list(mean_mass_g = 12)
) |>
  head(2)
```

The second call multiplies the predicted count by an external mean mass of
12 g, giving the posterior of expected biomass per site. Compatibility
limits scale proportionally because the constant has no uncertainty
attached.

## Group-level summaries with `mcmc_derive_data()`

`mcmc_derive_data()` pairs MCMC samples with the per-row data and returns an
`mcmc_data` object. `group_by()` + `summarise()` then aggregates across rows
within each group per MCMC iteration to produce per-group posterior distribution summaries.

### Total expected count per treatment

```{r mcd-total}
mcmc_derive_data(analysis, new_data = data, term = "^eCount$") |>
  group_by(treatment) |>
  summarise() |>
  coef()
```

The default summary function is `sum`, which returns the posterior of *total* abundance per treatment across all rows. 

### Per-site mean abundance
Provide a different summarise function, or define a custom one:
```{r mcd-per-site}
mcmc_derive_data(analysis, new_data = data, term = "^eCount$") |>
  group_by(site) |>
  summarise(.fun = mean) |>
  coef()
```

A coefficient of variation across sites measures spatial heterogeneity in
expected counts. 

```{r mcd-custom-fun}
mcmc_derive_data(analysis, new_data = data, term = "^eCount$") |>
  group_by(treatment) |>
  summarise(.fun = function(x) sd(x) / mean(x)) |>
  coef()
```

## Combining across analyses with `combine_samples()`

`mcmcr::combine_samples()` (with the `mcmc_data` method from `mcmcdata`)
joins two `mcmc_data` objects on shared data columns and applies a function
to align MCMC draws row-by-row. Its natural use is composing quantities
derived from *independent* analyses: for example, multiplying an abundance
posterior by a mass-per-fish posterior to get biomass, where each component
was fitted from a different dataset.

Assume a separate `mass_analysis` was fitted to size-measurement
data (different observation model, same site factor), and both analyses
expose a per-site quantity:

```{r combine-samples, eval = FALSE}
# Abundance posterior per site, from the counts model fitted above
abundance <- mcmc_derive_data(
  analysis,
  new_data = xnew_data(data, site),
  term = "^eCount$"
)

# Mass-per-fish posterior per site, from a separate analysis (sketch)
mass_per_fish <- mcmc_derive_data(
  mass_analysis,
  new_data = xnew_data(mass_data, site),
  term = "^eMass$"
)

# Per-site biomass posterior: abundance * mass-per-fish
biomass <- mcmcr::combine_samples(
  abundance,
  mass_per_fish,
  by = "site",
  fun = prod
)

coef(biomass)
```

`by` names the data columns to join on (here `site`); rows present in both
inputs are paired. `fun` is applied to the matched MCMC draws, so `prod`
gives the row-wise product across iterations. The result is an `mcmc_data`
with one row per joined key and a propagated posterior in `biomass`.

For the two analyses to combine:

* **MCMC dimensions must match.** Both fits need the same number of chains
  and iterations (e.g. both 3 chains × 500 saved draws). `combine_samples()`
  pairs draws position-by-position; mismatched dimensions error out.
* **`by` columns must exist in both `$data` slots.** Rows are paired by an
  inner join, so non-matching rows are dropped silently. Make sure factor
  levels and column types line up.

Within a single analysis, prefer expressing the composition directly in
`new_expr`; `combine_samples()` earns its keep when the components come
from independently fitted models whose draws need joining without re-running
them.

## Scalar derived quantities with `mcmc_derive()`

`mcmc_derive()` returns the raw MCMC samples for one or more scalar terms in
`new_expr`. This is useful when you want to compute functions of parameters with
full uncertainty propagation.

### Pull related scalars by regex

Both `eBaseCount` and `eRestoredEffect` are extracted in one call, sharing
the same iteration ordering:

```{r mcmc-derive-regex}
scalars <- mcmc_derive(
  analysis,
  new_data = character(0),
  term = "^(eBaseCount|eRestoredEffect)$"
)

coef(scalars)
```

### Arithmetic on `mcmcr` objects propagates uncertainty

Multiplying `mcmcr` objects acts element-wise across MCMC iterations:

```{r mcmc-derive-arithmetic}
eExpectedRestored <- scalars$eBaseCount * scalars$eRestoredEffect
coef(eExpectedRestored)
```

This is the absolute expected count at reference site and temperature under
the restored treatment: the analytic posterior of a function of parameters,
not a posterior over a per-row prediction grid.

## Random-effect zeroing

`random_effects` controls which `z_b*` parameters are eligible for
collapsing to the population level. A `z_b*` parameter is zeroed when its
associated factor in `new_data` is constant at level 1 (e.g. the factor
was omitted from `xnew_data()`).

### Default behaviour

`xnew_data(data, temperature)` leaves both `site` and `annual` at level 1,
so both `z_bSite` and `z_bSiteAnnual` are zeroed, giving the population-level
prediction:

```{r re-default}
xnew_data(data, temperature) |>
  predict(analysis, new_data = _) |>
  head(3)
```

### `random_effects = FALSE`

Skip all zeroing; the random effects are kept and `predict()` returns the
estimate for `site` level 1 and `annual` level 1 literally rather than the
population-level prediction:

```{r re-false}
xnew_data(data, temperature) |>
  predict(analysis, new_data = _, random_effects = FALSE) |>
  head(3)
```

### Named list: zero a subset of random effects

Narrow the candidate list to `z_bSite` only. `z_bSiteAnnual` is now kept
regardless, so the prediction uses level 1 of `annual` literally, while
`z_bSite` is still zeroed (because `site` is constant at level 1):

```{r re-named}
xnew_data(data, temperature) |>
  predict(
    analysis,
    new_data = _,
    random_effects = list(z_bSite = "site")
  ) |>
  head(3)
```

## Cheat sheet

| Goal | Tool |
|---|---|
| Posterior summary at new covariate values | `predict()` |
| Posterior summary for a scalar derived quantity | `predict()` with `new_data = character(0)` and `term` specified |
| Prediction for an unobserved factor level (includes amongst-level variation) | `new_expr` term with `rnorm(1, 0, s*)` for the random effect |
| Effect size (proportional, absolute) | `predict()` with `ref_data` + `ref_fun2` |
| Group-level aggregates (sum, mean, custom) per MCMC iteration | `mcmc_derive_data()` + `group_by()` + `summarise()` |
| Combine quantities derived from independent analyses | `mcmcr::combine_samples()` |
| Scalar parameter posteriors and arithmetic | `mcmc_derive()` |

See `?predict.mb_analysis`, `?mcmc_derive.mb_analysis`,
`?mcmc_derive_data.mb_analysis`, and `?mcmcr::combine_samples` for full
argument documentation.


